---
title: "Sampling biomass estimates from allometric equations"
format: 
  pdf:
    number-sections: true
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{newunicodechar}
        \newunicodechar{Ę»}{`} 
        \renewcommand{\thefigure}{A\arabic{figure}}
        \renewcommand{\thesection}{A\arabic{section}}
knitr:
  opts_chunk: 
    dev: cairo_pdf
crossref:
  fig-title: "Appendix Figure"
  fig-prefix: "Appendix Figure"
  tbl-title: "Appendix Table"
  tbl-prefix: "Appendix Table"
csl: ecology-letters.csl
bibliography: hawaii-arth-mete.bib
---

## Setup

We make use of these packages

```{r}
library(dplyr)
library(ggplot2)
library(cowplot)
```

## Allometric relationships

@gruner2003 published coefficients for taxon- and stage-specific allometric relationships of the form

$$
\text{mass} = a (\text{length})^b
$$

This allometric relationship was estimated from real data based on the log-log transform and assuming a normal distribution of log-transformed mass

$$
\begin{aligned}
\widehat{{\log(\text{mass})}} &= \log(a) + b \log(\text{length}) \\
\log(\text{mass}) &\sim \mathscr{N}(\widehat{{\log(\text{mass})}}, ~\sigma)
\end{aligned}
$$


@gruner2003 also published $R^2$ coefficients for these log-log transformed allometric relationships.

@gruner2007 measured length of individual specimens to 1 mm resolution and used the estiamted coefficients from these allometric relationships to approximate the mass of each specimen as $e^{\log(a) + b \log(\text{length})}$. This is a reasonable assumption but leads to potential issues when comparing data to Maximum Entropy Theory of Ecology [METE; @harte2011] predictions because METE assumes biomass-derived metabolic rates are continuous while using expected values based on 1 mm resolution length measurements introduces artificial discretization of the data.


## Drawing random biomass samples

To overcome this potential issue we conduct our METE analyses with random draws from $\log(\text{mass}) \sim \mathscr{N}(\widehat{{\log(\text{mass})}}, ~\sigma)$, one draw per individual. Random draws will be centered around $\widehat{{\log(\text{mass})}}$ but be continuous-valued rather than artificially discretized.

In order to achieve those random draws we must match the allometric coefficients from @gruner2003 to the taxonomy and stage information in @gruner2007. A modified version of Table 1 from @gruner2003 is shown in @tbl-grun

```{r}
#| label: tbl-grun
#| tbl-cap: |
#|     Coefficient estimates for the allometric relationship 
#|     $\text{mass} = a (\text{length})^b$, modified from Table 1 of 
#|     @gruner2003

allom <- read.csv("../gruner2003.csv")

knitr::kable(allom)
```


Now we match rows in `allom` to rows in the data from @gruner2007 (which we call `arth`) based on taxonomy and stage. Matching follows a priority hierarchy from genus to class and from specific stage to all stages lumped.


```{r}
#| label: match-arth

arth <- read.csv("../gruner2007.csv")

#' function to match based on taxonomy and stage, prioritizing finest 
#' resolution match possible. to be applied over rows of `arth` with `mapply`.
#' a few key assumptions are made:
#'     - `arth$stage` has no missing data (it does not)
#'     - `allom[, c("taxon", "stage")]` has no duplicates (it does not)
#'
#' @param genus genus column from `arth`
#' @param family family column from `arth`
#' @param order order column from `arth`
#' @param class class column from `arth`
#' @param stage stage column from `arth`
#' @param d data frame of allometric coefficients
#' 
#' @returns integer ID of row in allometry data frame 

match_allom <- function(genus, family, order, class, stage, d) {
    tax_levels <- c("genus", "family", "order", "class")
    
    # vector holding specific taxon name at each level
    taxvals <- c(genus, family, order, class) |> 
        setNames(tax_levels)

    tmatch <- taxvals %in% d$taxon
    
    # if nothing matches, return arthropod all
    if(sum(tmatch) == 0) {
        d_idx <- d$taxon == "Arthropoda" & d$stage == "all"
    } else {
        
        t_idx <- which(tmatch) |> min()
        tax <- taxvals[t_idx]
        
        d_idx <- d$taxon == tax
        
        # multiple rows match so match stage
        if(sum(d_idx) > 1) {
            d_idx <- d_idx & (stage == d$stage)
        } 
    }
    
    return(which(d_idx))
}


# apply to every row of `arth`
allom_idx <- mapply(match_allom,
                    arth$genus, arth$family, arth$order, 
                    arth$class, arth$stage,
                    MoreArgs = list(d = allom))

# add allom info to `arth` data
arth$taxon_matched <- allom$taxon[allom_idx]
arth$stage_matched <- allom$stage[allom_idx]
arth$a <- allom$a[allom_idx]
arth$b <- allom$b[allom_idx]
arth$R2 <- allom$R2[allom_idx]

# add prediction to `arth`
arth$biom_pred <- arth$a * arth$lengthmm^arth$b

# where prediction is NA (due to missing length) fill in with estimate from
# Gruner 2007
arth$biom_pred[is.na(arth$biom_pred)] <- arth$ind_biom[is.na(arth$biom_pred)]
```

Now we are ready to draw biommass samples.


## Draw log(biomass) using considering $R^2$

To draw a sample of biomass we have to know $\sigma$, i.e. the standard deviation of the residuals of the log-log allometry model. We can back-calculate $\sigma$ from $R^2$ using sums of squares:

$$
\begin{aligned}
R^2 &= \frac{SS_\text{regression}}{SS_\text{regression} + SS_\text{residual}}
\\ 
\Rightarrow SS_\text{residual} &= SS_\text{regression} \cdot \frac{1 - R^2}{R^2} 
\end{aligned}
$$

The residual standard deviation is connected to $SS_\text{residuals}$ by $\sigma = \sqrt{SS_\text{residual} / (n - 2)}$ where $n$ is number of specimens. To calculate $SS_\text{regression}$ and $n$ we have to group the `arth` data according to which taxon and stage each row matched to


```{r}
#| label: sigma-resid

# record predicted log biomass for further calculations
log_biom_pred <- log(arth$biom_pred)

# vector to hold residual standard deviations
sigma_resid <- numeric(length(log_biom_pred))

# group by matching index
groups <- split(1:nrow(arth), allom_idx)

# loop over groups calculating n and sigma
for(g in names(groups)) {
    # rows in `arth` to consider
    idx   <- groups[[g]]
    
    # predictions for these groups
    preds <- log_biom_pred[idx]
    
    # number specimens for this group
    n_g <- length(idx)

    if(n_g < 3) {
        # too few points to estimate variance, assume sigma = 0
        s <- 0
    } else {
        # sum of squares regression
        SS_reg <- sum((preds - mean(preds))^2)
        
        if(SS_reg == 0) {
            # occurs when all length values are equal or slope = 0,
            # assume sigma = 0
            s <- 0
        } else {
            # calculate sigma of residuals
            R2_g   <- allom$R2[as.integer(g)]
            SS_res <- SS_reg * (1 - R2_g) / R2_g
            
            s <- sqrt(SS_res / (n_g - 2))
        }
    }
    
    # update sigma vector
    sigma_resid[idx] <- s
}

# add sigma of residuals to `arth`
arth$sigma_resid <- sigma_resid

```

Finally a random draw of biomass can be added to `arth` by sampling from

$$
\log(y) \sim \mathscr{N}(\log(\text{predicted biomass}), ~\sigma)
$$


```{r}
#| label: draw-biom

# set seed for consistent results
set.seed(123)

arth$biom_sim <- exp(rnorm(nrow(arth), log(arth$biom_pred), arth$sigma_resid))

```

## Sanity check against existing `ind_biom`

The published biomass estimates from @gruner2007 are retained in `arth$ind_biom`. If our random samples of biomass were successful then `(biom_sim - ind_biom) / sigma_resid` should follow a normal distribution with mean = 0 and standard deviation = 1.

```{r}
# create a new data frame with a column that should be N(0, 1)
normalized <- filter(arth, sigma_resid > 0) |> 
    mutate(norm_01 = (biom_sim - ind_biom) / sigma_resid) 
```

A few specimens ($n =$ `r sum(normalized$norm_01 > 5 | normalized$norm_01 < -5)` out of `r nrow(arth)` total specimens) are outliers. The most reliable treatment for those outliers is to set them equal to their originally published biomass from @gruner2007

```{r}
# find problem specimens
trouble <- filter(normalized, norm_01 > 4) |> 
    select(specimen_code) |> 
    pull() |> 
    (\(x) arth$specimen_code %in% x)()

# set biomasses to original published
arth$biom_pred[trouble] <- arth$ind_biom[trouble]
arth$biom_sim[trouble] <- arth$ind_biom[trouble]

# ensure any future sampling will respect the original published value
arth$sigma_resid[trouble] <- 0

```

Now we can re-calculate the normalized samples and compare to $\mathscr{N}(0, ~1)$ in @fig-check.

```{r}
#| label: fig-check
#| fig-cap: |
#|     Checking normalized sample of biomass against expectation of $\mathscr{N}(0, ~1) shown in red$
filter(arth, sigma_resid > 0) |> 
    mutate(norm_01 = (log(biom_sim) - log(ind_biom)) / sigma_resid) |> 
    ggplot(aes(norm_01)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50) +
    stat_function(fun = dnorm, color = "red", linewidth = 1) +
    xlab("Normalized biomass sample") +
    theme_cowplot() 

ggplot(arth, aes(ind_biom, biom_sim)) +
    geom_point() +
    scale_x_log10() +
    scale_y_log10()
```

The random draws indeed conform to the expectation.

## Saving the final product

The final computation is to add relative metabolic rate which we assume is simply

$$
\text{metabolic rate} \propto \text{biomass}^{3/4}
$$

METE normalizes metabolic rate such that the minimum is always 1, thus we only need to raise biomass to the $3/4$ power

```{r}
arth$metabolic_rate <- arth$biom_sim^0.75

```

We ship these data with the R package *HawaiiArthMETE* so our last step is to save them to the `/data` directory.

```{r}
save(arth, file = "../../data/arth.rda")
```


## References
