---
title: "Eco-evolutionary deviations from steady state across the Hawaiian chronosequence from a maximum entropy perspective"
author:
  - name: Andrew J. Rominger
  - name: Kaite Thai
  - name: Rosemary G. Gillespie
  - name: John Harte
  - name: Daniel S. Gruner
format:
  pdf:
    keep-tex: true
    include-in-header: 
      text: |
        \usepackage{newunicodechar}
        \newunicodechar{ʻ}{`} 
        \usepackage{xr}
        \externaldocument{hawaii-arth-mete-supp}
        \newcommand{\sfig}[1]{Supplementary Figure~\ref{#1}}
        \newcommand{\ssec}[1]{Supplementary Section~\ref{#1}}
knitr:
  opts_chunk: 
    dev: cairo_pdf
csl: ecology-letters.csl
bibliography: hawaii-arth-mete.bib
---


```{r}
#| label: setup
#| include: false


library(ggplot2)
library(cowplot)
library(grid)
library(sf)
library(dplyr)
library(meteR)
library(HawaiiArthMETE)

knitr::opts_chunk$set(fig.width = 5.5, fig.height = 4, echo = FALSE)

```

## Introduction

Biodiversity patterns are bound by two seemingly contradictory states: they show striking consistency in form across systems, and yet also must emerge in spite of, or because of, the constant flux and dynamic change of ecology and evolution. The near universality of biodiversity patterns includes the scaling of species richness with area, the ubiquity of rare spaces and scarcity of common species, and the highly skewed distribution of biomass or metabolic resource use across organisms. Debate has persisted for over a century as to whether these consistent patterns form because of ecological and evolutionary drivers that force systems into these configurations, or are largely the outcome of chance acting on large systems of constituent parts. The former gives rise to niche-based theories of coexistence, while that later has produced a number of neutral-like models.

Both deterministic and stochastic perspectives on the generation of universal biodiversity patterns most often make the choice of considering an ecosystem at "equilibrium" or steady state. However, ecosystems are known to be dynamic both through intrinsic eco-evolutionary processes and due to recent anthropogenic perturbation. Whether universal patterns are insensitive to non-steady-state processes or are a result of those processes remains addressed. To better understand the interaction of non-steady-state processes and universal patterns, we combine data from the highly dynamic ecosystems of the Hawaiian Archipelago with the intentionally minimalist steady state predictions of the Maximum Entropy Theory of Ecology.

### The Hawaiian Chronosequence 

- world famous adaptive ration
- arthropods as diverse and useful study system
- age differences are snapshots in eco-evolutionary process


```{r}
#| label: fig-chrono-work
#| include: false

x <- st_read("../hi_geol/hawaii_state_geol_ageClean.shp") |> 
    filter(!is.na(age_mid)) 
hi <- st_read("../hi_geol/hi_outline.shp")
sites <- st_read("../hi_geol/gruner_sites.kml") |> 
    st_transform(crs = st_crs(x))

p <- ggplot(hi) +
    geom_sf(color = "gray30", fill = "gray30", linewidth = 0.1) +
    geom_sf(data = x,
            aes(fill = age_mid, color = age_mid),
            linewidth = 0.0001,
            inherit.aes = FALSE) +
    scale_fill_viridis_c(option = "magma", begin = 0.2, end = 1,
                         transform = c("log10", "reverse")) +
    scale_color_viridis_c(option = "magma", begin = 0.2, end = 1,
                         transform = c("log10", "reverse")) +
    geom_sf(data = sites, shape = 21, fill = "black", color = "white") +
    theme_cowplot() + 
    theme(legend.position = "none", 
          axis.line = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())


log_age_range <- range(log(x$age_mid, 10))
log_age <- pretty(log_age_range) |> 
    (\(x) x[x > log_age_range[1] & x < log_age_range[2]])()


l <- data.frame(y = seq(log_age_range[1], 
                   log_age_range[2], 
                   length.out = 500), 
           z = 1) |> 
    mutate(u = 10^y) |>
    ggplot(aes(z, y, fill = u, color = u)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", begin = 0.2, end = 1, 
                         transform = c("log10", "reverse")) +
    scale_color_viridis_c(option = "magma", begin = 0.2, end = 1, 
                         transform = c("log10", "reverse")) +
    annotate("segment",
        x = rep(c(0.47, 1.53), each = length(log_age)),
        xend = rep(c(0.7, 1.3), each = length(log_age)),
        y = rep(log_age, 2),
        yend = rep(log_age, 2), 
        color = "white", linewidth = 0.5) +
    annotate("segment",
        x = rep(-0.1, length(unique(arth$flow_age_my))),
        xend = rep(0.49, length(unique(arth$flow_age_my))),
        y = log(unique(arth$flow_age_my), 10),
        yend = log(unique(arth$flow_age_my), 10), 
        arrow = arrow(length = unit(0.25, "npc"))) +
    annotate("text", 
             x = 0.75, y = 1.5, 
             label = "Age (mya)") +
    scale_x_continuous(expand = FALSE) +
    scale_y_continuous(
        expand = FALSE,
        limits = c(-10, 5),
        sec.axis = sec_axis(
            ~ ., 
            breaks = rev(log_age),
            labels = sapply(rev(log_age), function(a) bquote(10^.(a)))
        )
    ) +
    theme_cowplot() +
    theme(legend.position = "none",
          axis.ticks = element_blank(), 
          axis.line = element_blank(), 
          axis.title.y = element_blank(), 
          axis.title.x.bottom = element_blank(),
          axis.text.x = element_blank(), 
          axis.text.y.left = element_blank(), 
          axis.text.y.right = element_text(size = 10)) + 
    coord_cartesian(clip = "off")

```

```{r}
#| label: fig-chrono
#| cache: true
#| fig-cap: |
#|     Chronosequence of the Hawaiian Archipelago showing oldest flows to the
#|     west on Kauaʻi and youngest flows to the east on Hawaiʻi. The five 
#|     sites are indicated with dots on the map and arrows indicating their
#|     flow ages on the legend. From oldest to youngest (west to east) sites 
#|     are: Alakaʻi, Kamakou, Kohala, Laupāhoehoe, Kīlauea.

plot_grid(p, l, rel_widths = c(1, 0.2))
```


### The Maximum Entropy Theory of Ecology

- METE posits: if state variables are in steady state, all biodiversity patterns are statistical outcomes from those macroscopic variables
- departures from METE represent departures from steady state

We find departures from steady state at middle aged sites. Departures are predicted by community dissimilarity which peaks at middle aged sites. This suggests these communities are unstable, perhaps due to their age being the right time for adaptive radiation to take off for arthropods. More work connecting departure from eco steady state and evo steady state would be needed to further test this possibility.


## Methods

### Reproducibility and data preparation

A custom R package, *HawaiiArthMETE* [@HawaiiArthMETE], accompanies this paper and can be used to reproduce all results. All data analyzed here are available via the package as described in @sec-setup. 

Arthropod data come from the original publication by @gruner2007 and include `r nrow(arth)` specimens. Each specimen was identified to the species level (morpho species were used in `r round(100 * mean(grepl("sp_", arth$species)), 2)`% of specimens). Specimen lengths were measured to the nearest millimeter.  For METE analyses we need to estimate metabolic rate from these data. Because METE normalizes metabolic rates such that the lowest possible rate is 1 we use a simplified $3/4$ scaling rule to approximate metabolic rate $B$ from mass $M$ as $B \propto M^{3/4}$.

This leaves the estimation of mass from length. @gruner2007 used taxon- and stage- specific length-to-biomass allometric scaling relationships [@gruner2003] to approximate mass as the expected value of a log-log regression

$$
\begin{aligned}
\text{mass} & = e^{\widehat{\log(\text{mass})}} \\
\widehat{\log(\text{mass})} &= \log(a) + b \cdot \log(\text{length})
\end{aligned}
$$

where $a$ and $b$ are estimated from data collected by @gruner2003 and vary by taxon and stage.

Using these expected value masses based on length measured to the nearest milliliter yields metabolic rate approximations that are artificially discretization (see the appendix "Sampling biomass estimates from allometric equations"). These descretizations in turn lead to spurious conclusions about the predicitiveness of METE because METE assumes metabolic rates are continuous (see @sec-re-run).

Therefore, instead of estimating mass as the expected value from an allometric relationship we sample masses as random draws from the full model relating mass to length

$$
\begin{aligned}
\text{mass} & = e^{y} \\
y &\sim \mathscr{N}(\widehat{\log(\text{mass})}, ~\sigma) \\
\widehat{\log(\text{mass})} &= \log(a) + b \cdot \log(\text{length})
\end{aligned}
$$

where $\sigma$ is the taxon- and stage-specific standard deviation of residuals. The full procedure for sampling masses is detailed in the appendix "Sampling biomass estimates from allometric equations" and the consequences of using random draws versus expected values on the METE analysis is thoroughly investigated in @sec-re-run. We conclude that using random draws for mass estimates is largely similar to using expected values, and our overall conclusions are unchanged, but using draws avoids both spurious deviations from and conformations to METE that would be introduced by using expected values of mass.

### METE analysis

To produce METE predictions for the SAD and IPD we use the *meteR* R package [@rominger-merow2016]. We compare these predictions to observed SAD and IPD patterns using a $z^2$-value. The use of $z^2$-values is discussed in @rominger-merow2016, but briefly: the $z^2$-value compares the log likelihood of the data given them METE model to the log likelihood of a hypothetical data set drawn from the METE model itself. This comparison is done many times and the actual $z^2$-value is calculated as

$$
z^2 = \left(\frac{\log(\mathscr{L}(\text{data} \mid \text{METE})) - \hat{\mu}_{\log\mathscr{L}_\text{METE}}}{\hat{\sigma}_{\log\mathscr{L}_\text{METE}}}\right)^2
$$

where $\hat{\mu}_{\log\mathscr{L}_\text{METE}}$ is the mean of the log likelihood of data drawn from METE itself, and $\hat{\sigma}_{\log\mathscr{L}_\text{METE}}$ is the standard deviation of the log likelihood of data drawn from METE itself.

If METE is truly correct in describing the data, $z^2$ will asymptotically have a $\chi^2$ distribution with $d.f. = 1$ [@rominger-merow2016]. If METE does not truly describe the data, the $z^2$-value is likely to be large.

We also quantify the difference between METE predictions and observed SADs and IPDs by comparing the tails of the observed and predicted distributions. These tails can be especially informative about the nature of deviations [@harte2011]. For the SAD we compare rarity as the log ratio of observed proportion of singletons to the predicted proportion of singletons, and we compare dominance as log ration of observed total abundance of the top three most abundant species to the predicted total abundance of the top three most abundant species. For the IPD we compare metrics similar to rarity and dominance: the log ratio of observed proportion of individuals with a scaled metabolism of less than 3 to its prediction, and the log ratio of the observed sum of metabolic rates for the top 10 individuals to its prediction.



## Statistical analysis of deviations from METE and potential explanatory variables

We first evaluate how deviations from METE, as measured by $z^2$-values, vary across the chronosequence. We then test for relationships between $z^2$-values and proportion of non-native species at a site as well as within site, between sample dissimilarity or $\beta$-diversity.

For all these analyses we fit linear models predicting log-transformed $z^2$-values responding to each explanatory variable: substrate age, proportion non-native species, and within site $\beta$-diversity. Log-transforming $z^2$-values brings them closer to the normal assumptions of a linear model. The need for log-transformation is not surprising: while $z^2$-values will be $\chi^2$ distributed if METE is true, deviations from METE add more probability to larger values, and taken together will lead to right-skewed distribution of $z^2$-values. Log-transformation brings right-skewed random variables closer to normal.

When modeling deviations from METE, we have two statistical challenges to address: samples within sites could represent pseudo replicates necessitating a random-effects model; however, because explanatory variables are site-level characteristics, site-level random effects would be completely collinear with site-level explanatory variables. 

To address these challenges we model the site-level mean and variance rather than each individual sample within each site. This is the most conservative approach but runs the risk of being under-powered because while we have `r nrow(unique(arth[, c("site", "tree")]))` samples, we only have `r n_distinct(arth$site)` sites. Modeling mean and variance is the same technique used in meta-analyses. We therefore make use of the *metafor* package [@metafor] in fitting testing these models.

## Calculating $\beta$-diversity

To quantify within site, between sample $\beta$-diversity we first calculate mean Bray-Curtis dissimilarity between samples and then, to standardize $\beta$-diversity measures across sites, we normalize mean dissimilarity to a null model. We use the `quasiswap_count` algorithm implemented in *vegan* [@vegan] to generate null samples. This algorithm preserves row and column sums of a sample by species matrix and also seeks to maintain null matrix fill close to the observed fill thereby mimicking patchiness in the real data. We use 500 null sample by species matrices to generate null models for each site. Then normalized within site, between sample $\beta$-diversity was calculated as a z-score

$$
\text{Normalized }\beta = \frac{\text{BC}_\text{obs} - \mu_\text{null}}{\sigma_\text{null}}
$$

where $BC_\text{obs}$ is the observed mean Bray-Curtis dissimilarity between samples within a site, $\mu_\text{null}$ is the mean of the null distribution, and $\sigma_\text{null}$ is the standard deviation of the null distribution.

We are also interested in understanding how native and non-native species contribute to the overall mean dissimilarity at a site. We can do this by partitioning Bray Curtis dissimilarity between native and non-native groups of species. @sec-part provides the mathematical and computational details of how we do this. The end result of partitioning Bray-Curtis dissimilarity between native and non-native species is that we have a ratio of within site $\beta$-diversity attributable to native species to that attributable to non-native species. We take the log of this ratio such that equal contributions of native and non-native species yields a log ratio of 0.

We also compare this Bray Curtis dissimilarity partitioning to a null model. Here the `quasiswap_count` algorithm will not work because it permutes individuals across species as well as sites. Instead we want to only shuffle the label of each species: native or non-native. But we still need to maintain, at least probabilistically, attributes of the data that might confound null model analysis. Specifically, we should seek to maintain, in probability, the distribution of abundances of native and non-native species. We wrote a custom algorithm to do this in the *HawaiiArthMETE* package and described in @sec-bart.

## Results

```{r}
#| label: supp-read
#| include: false

ff <- list.files("hawaii-arth-mete-supp_cache/pdf/", pattern = ".RData", 
                 full.names = TRUE) |> 
    gsub(".RData", "", x = _)

for(f in ff) {
    lazyLoad(f)
}
```

### Deviations from METE across the chronosequence

```{r}
#| label: fig-z2-chrono
#| fig-cap: |
#|     Deviations from METE across the chronosequence as measured by 
#|     $z^2$-values. Colors represent sites and match the color coding of 
#|     substrate age in @fig-chrono.

fig_z2_chrono

```

- deviations for SAD and IPD peak at middle age
- deviations for IPD larger overall
- deviations of SAD are because of more unevenness (more singletons and more hyper abundant species)
- no consistent relationship between IDP deviations and the extremes of the distribution (deviation is in the middle)
- deviations are predicted by proportion of non-natives
- deviations are predicted by dissimilarity
- something is going on at Kīlauea


## Discussion

- dissimilarity as instability
- possible causes of instability---adaptive radiation
- presents future hypothesis to test with genetic data
- hypotheses
    - diversification drives divergence; diversification slowed down on older flows
    - invasion drives divergence
    - both do and they interact, perhaps even conditions favoring diversification favor invasion too
    
## References
