---
title: "Analysis of arthopod communites in Hawaii using METE Theory"
format: 
    pdf:
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
---

## Setup

```{r}
#| label: knitr-setup
#| include: false

knitr::opts_chunk$set(fig.width = 4.5, fig.height = 3)
```

```{r}
library(meteR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vegan)
library(metafor)
library(HawaiiArthMETE)

# load data 
data(arth)
```


## METE calculations


```{r}
#| label: mete-calc
#| message: false
#| output: false
#| cache: true

arth_mete <- split(arth, arth[, c("site", "tree")], drop = TRUE) |> 
    lapply(function(x) {
        # the esf for this grouping
        esf <- meteESF(spp = x$species_code, 
                       abund = x$abundance, 
                       power = x$metabolic_rate)
        
        # sad and ipd
        s <- sad(esf)
        p <- ipd(esf)
        
        data.frame(
            # sample info
            site = x$site[1],
            site_name = x$site_name[1],
            tree = x$tree[1],
            flow_age_my = x$flow_age_my[1],
            
            # z2 values
            z2_sad = logLikZ(s)$z,
            z2_ipd = logLikZ(p)$z,
            
            # prop non-native for abundance and power
            prop_n_non_nat = sum(
                x$abundance[!(x$origin %in% c("endemic", "indigneous"))]
            ) / 
                sum(x$abundance),
            prop_p_non_nat = sum(
                x$metabolic_rate[!(x$origin %in% c("endemic", "indigneous"))]
            ) / 
                sum(x$metabolic_rate),
            
            # differences in extreme values
            n1_diff = mean(meteDist2Rank(s) == 1) - mean(s$data == 1),
            nmax_diff = sum(meteDist2Rank(s)[1:3]) - sum(s$data[1:3]),
            p1_diff = mean(meteDist2Rank(p) <= 2) - mean(p$data <= 2),
            pmax_diff = sum(meteDist2Rank(p)[1:3]) - sum(p$data[1:3])
        )
    })

arth_mete <- do.call(rbind, arth_mete)




```


### What are the highly invaded, highly dominated samples?

```{r}
outliers <- filter(arth_mete, nmax_diff < -300)

```

Two samples stand out as clear outliers both in terms of deviation from SAD predictions and being the most highly invaded. The two samples come from the Laupāhoehoe and Kohala sites. To further understand these two samples and determine if they should remain in the analyses we look at their composition. @tbl-out1 shows the three most abundant species for the Laupāhoehoe outlier sample while @tbl-out2 shows the three most abundant species for the `r Kohala outlier sample.


```{r}
#| label: tbl-out1

filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(
        caption = paste(
            "Top three most abundant taxa from", 
            levels(arth$site_name)[as.numeric(outliers$site[1])], 
            "outlier sample"
        )
    )

```

```{r}
#| label: tbl-out2

filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(
        caption = paste(
            "Top three most abundant taxa from", 
            levels(arth$site_name)[as.numeric(outliers$site[2])], 
            "outlier sample"
        )
    )

```

In both outliers the most abundant species is non-native, and in the case of Laupāhoehoe, all top three most abundant species are non-native. Laupāhoehoe and Kohala share *Salina celebensis* in common which is a Collembola known to reach high densities in Hawaiʻi [@christiansen1992]. The abundances of *Salina celebensis* and *Greenidea formosana* in these samples are significant outliers compared to the maximum abundance found in any other sample as shown in @fig-crazy-abund.


```{r}
#| label: fig-crazy-abund

crazy_abund <- filter(arth, 
                      (site %in% outliers$site & tree %in% outliers$tree)) |> 
    summarize(n = n(), .by = c(site, tree, species_code)) |> 
    arrange(desc(n)) |> 
    head(3)

summarize(arth, .by = c(site, tree), 
          nmax = max(table(species_code))) |> 
    filter(nmax < 300) |> 
    ggplot(aes(nmax)) +
    geom_histogram(bins = 20) +
    geom_segment(data = crazy_abund, 
                 mapping = aes(x = n, y = rep(10, 3), 
                               xend = n, yend = rep(0, 3)), 
                 color = "red", 
                 arrow = arrow(length = unit(1, "char")),
                 inherit.aes = FALSE) +
    xlab("Maximum abundances") +
    ylab("Count") +
    this_ms_look()




```

Because these samples are so unusual we decided to drop them from all further analyses presented in the main text. However, we retain the outlier samples in the supplement to validate that our results are not biased by their removal.

```{r}
#| label: remove-outliers

# remove from mete results
arth_mete_out <- arth_mete
arth_mete <- filter(arth_mete, 
                    !((site %in% outliers$site) & (tree %in% outliers$tree)))

# remove from arth data frame
arth_out <- arth
arth <- filter(arth, 
               !((site %in% outliers$site) & (tree %in% outliers$tree)))

```




## Deviations from METE across the chronosequence

```{r}
#| fig-width: 4.5
#| fig-height: 3

ggplot(arth_mete, aes(flow_age_my, z2_sad, color = site_name)) +
    geom_jitter(width = 0.1) +
    this_ms_look(log_x = TRUE, log_y = TRUE) 



ggplot(arth_mete, aes(flow_age_my, z2_ipd, color = site_name)) +
    geom_jitter(width = 0.1) +
    scale_x_log10() +
    scale_y_log10() + 
    scale_color_viridis_d(option = "magma") 

```

## How do observed patterns deviate from METE?

@fig-sad-how shows that when METE predictions deviate from observed SADs, METE under-predicts both rarity and dominance compared to observed SADs. This means that observed SADs are more uneven than the METE prediction. 


```{r}

ggplot(arth_mete, aes(z2_sad, z2_ipd, color = site)) +
    geom_point() +
    scale_x_log10(limits = range(select(arth_mete, starts_with("z2")))) +
    scale_y_log10(limits = range(select(arth_mete, starts_with("z2")))) +
    scale_color_viridis_d(option = "magma") +
    geom_abline(slope = 1, intercept = 0)


ggplot(arth_mete, aes(z2_sad, n1_diff, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(nmax_diff, z2_sad, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(p1_diff, z2_ipd, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(pmax_diff, z2_ipd, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma") 



```

## How do METE and the real data differ

Which one over-predicts rarity?

## Comparing deviations from METE with potential explanatory variables



### Deviation vs. proportion that are invasive

```{r}
#| label: fig-dev-non


ggplot(arth_mete, aes(prop_n_non_nat, z2_sad, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma")


ggplot(arth_mete, aes(prop_p_non_nat, z2_ipd, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma")




```



### Deviation vs. dissimilarity

```{r}
#| eval: false

z_diss <- function(x, B = 500) {
    obs_dist <- x |> vegdist() |> mean()
    
    sim <- nullmodel(x, "quasiswap_count") |>
        simulate(nsim = B)

    sim_mean_dist <- sapply(1:dim(sim)[3], function(i) {
        vegdist(sim[, , i]) |> mean()
    })


    (obs_dist - mean(sim_mean_dist)) / sd(sim_mean_dist)
    
}


diss <- split(arth, arth$site) |> 
    sapply(FUN = function(dat) {
        summarize(dat, abund = n(), .by = c(tree, species_code)) |> 
            pivot_wider(names_from = species_code, 
                        values_from = abund, 
                        values_fill = 0) |> 
            select(!tree) |> 
            z_diss() 
    })

mete_by_diss <- mutate(arth_mete, 
                       mean_diss = diss[site])



ggplot(mete_by_diss, aes(mean_diss, z2_sad, color = site)) +
    geom_jitter(width = 0.25) +
    this_ms_look(log_y = TRUE)


xy <- mete_by_diss |> 
    summarize(z2_sad_mean = mean(log(z2_sad)), 
              z2_sad_sd = sd(log(z2_sad)), 
              n = n(), 
              mean_diss = first(mean_diss), 
              .by = site)



rma(yi = z2_sad_mean, 
    vi = z2_sad_sd^2 / n, 
    mods = ~ mean_diss, 
    data = xy, 
    method = "REML")


```


