---
title: "Supplement to \"Eco-evolutionary deviations from steady state across the Hawaiian chronosequence from a maximum entropy perspective\""
format: 
    pdf:
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
knitr:
    opts_chunk: 
        dev: cairo_pdf
---


This supplement walks through the steps of recreating the entire analysis presented in the main text. It also provides further analysis in support of and to contextualize results in the main text. 


## Computing setup

Analyses are supported by a custom R package *HawaiiArthMETE* [@foo] which can be installed from github. 

```{r}
#| label: pkg-install
#| eval: false

devtools::install_github("ecoevomatics/HawaiiArthMETE")
```

The quarto [@quarto] document used to generate this supplement can be found with the R command `system.file("ms/hawaii-arth-mete-supp.qmd", package = "HawaiiArthMETE")`.

The following R computing environment is used for all analyses:

```{r}
#| label: setup

library(meteR)
library(ggplot2)
library(ggh4x)
library(dplyr)
library(tidyr)
library(HawaiiArthMETE)


knitr::opts_chunk$set(
    # default figure dims
    fig.width = 5.5, fig.height = 4, 
    # caching is default for all subsequent chunks
    cache = TRUE
)

# load data 
data(arth)

sessionInfo() |> 
    print(locale = FALSE)
```


## METE calculations

Comparison of the observed data to predictions from the Maximum Entropy Theory of Ecology (METE) is performed with package *meteR* [@rominger-merow2016]. The following code records information about each sample (name, flow age, proportion non-native presence) as well as several metrics of deviation between theory and observation:

- a $z^2$ value measuring a normalized difference in likelihoods between observed data and theoretical expectation
- differences in proportion of rarity (number of singletons or normalized metabolic rate $\leq 2$) observed and expected
- differences in proportion of dominance (sum of top three most abundant species or sum of top three highest normalized metabolic rates) observed and expected

All metrics are calculated for both the species abundance distribution (SAD) and individual metabolic rate (or power) distribution (IPD).

```{r}
#| label: mete-calc
#| message: false
#| output: false
#| cache: true

arth_mete <- split(arth, arth[, c("site", "tree")], drop = TRUE) |> 
    lapply(function(x) {
        # the esf for this grouping
        esf <- meteESF(spp = x$species_code, 
                       abund = x$abundance, 
                       power = x$metabolic_rate)
        
        # sad and ipd
        s <- sad(esf)
        p <- ipd(esf)
        
        data.frame(
            # sample info
            site = x$site[1],
            site_name = x$site_name[1],
            tree = x$tree[1],
            flow_age_my = x$flow_age_my[1],
            
            # z2 values
            z2_sad = logLikZ(s)$z,
            z2_ipd = logLikZ(p)$z,
            
            # differences in extreme values
            n1_diff = log(mean(meteDist2Rank(s) == 1) / mean(s$data == 1)),
            nmax_diff = log(sum(meteDist2Rank(s)[1:10]) / sum(s$data[1:10])),
            p1_diff = log(mean(meteDist2Rank(p) <= 3) / mean(p$data <= 3)),
            pmax_diff = log(sum(meteDist2Rank(p)[1:10]) / sum(p$data[1:10]))
        )
    })

arth_mete <- do.call(rbind, arth_mete)




```

<!-- An initial investigation of patterns in the SAD reveal two clear outliers in terms of deviation from theory and proportion of non-native individuals (@fig-z2-out).  -->


<!-- ### Investigating highly invaded, highly dominated samples -->

<!-- The two outlier samples both show a difference (theory - observed) in maximum abundance of over -300 (@fig-z2-out).  -->

```{r}
#| label: calc-outliers
#| dependson: mete-calc
#| eval: false

outliers <- filter(arth_mete, nmax_diff < -300)
```

```{r}
#| label: fig-z2-out
#| fig-width: 7
#| dependson: [mete-calc]
#| eval: false

names(mete_by_non)
names(arth_mete)
arth_mete <- full_join(arth_mete, mete_by_non, by = c("site", "tree"))

select(arth_mete, site_name.x, prop_nspp_non_nat, nmax_diff.x, z2_sad.x) |> 
    # pivot so both metrics are plotted in faceted panels
    pivot_longer(prop_nspp_non_nat:nmax_diff.x, names_to = "metric") |> 
    ggplot(aes(value, z2_sad.x, color = site_name.x)) +
    geom_point() +
    
    # add arrows pointing to outliers
    geom_segment(
        data = pivot_longer(outliers, 
                            c(prop_n_non_nat, nmax_diff), 
                            names_to = "metric"), 
        aes(x = value * 0.8, y = z2_sad * 0.9, 
            xend = value * 0.97, yend = z2_sad * 0.99), 
        arrow = arrow(length = unit(1, "char")),
        inherit.aes = FALSE
    ) +
    
    # facet by metric and trick strip labels into being
    # axis labels
    facet_grid(
        cols = vars(metric), scales = "free_x", 
        labeller = as_labeller(
            c(nmax_diff.x = "Diff. in maximum abundance", 
              prop_nspp_non_nat = "Prop. non-native individuals")
        ), 
        switch = "x"
    ) +
    ylab(expression("SAD"~z^2*"-value")) +
    this_ms_look() +
    theme(strip.background = element_blank(), 
          strip.placement = "outside", 
          axis.title.x = element_blank())
```

<!-- These two outlier samples come from the Laupāhoehoe and Kohala sites. To further understand these two samples and determine if they should remain in the analyses we look at their composition. @tbl-out1 shows the three most abundant species for the Laupāhoehoe outlier sample while @tbl-out2 shows the three most abundant species for the `r Kohala outlier sample. -->


```{r}
#| label: tbl-out1
#| dependson: [mete-calc]
#| eval: false

filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(
        caption = paste(
            "Top three most abundant taxa from", 
            levels(arth$site_name)[as.numeric(outliers$site[1])], 
            "outlier sample"
        )
    )

```

```{r}
#| label: tbl-out2
#| dependson: [mete-calc]
#| eval: false

filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(
        caption = paste(
            "Top three most abundant taxa from", 
            levels(arth$site_name)[as.numeric(outliers$site[2])], 
            "outlier sample"
        )
    )

```

<!-- In both outliers the most abundant species is non-native, and in the case of Laupāhoehoe, all top three most abundant species are non-native. Laupāhoehoe and Kohala share *Salina celebensis* in common which is a Collembola known to reach high densities in Hawaiʻi [@christiansen1992]. The abundances of *Salina celebensis* and *Greenidea formosana* in these samples are significant outliers compared to the maximum abundance found in any other sample as shown in @fig-crazy-abund. -->


```{r}
#| label: fig-crazy-abund
#| dependson: [mete-calc]
#| eval: false

crazy_abund <- filter(arth, 
                      (site %in% outliers$site & tree %in% outliers$tree)) |> 
    summarize(n = n(), .by = c(site, tree, species_code)) |> 
    arrange(desc(n)) |> 
    head(3)

summarize(arth, .by = c(site, tree), 
          nmax = max(table(species_code))) |> 
    filter(nmax < 300) |> 
    ggplot(aes(nmax)) +
    geom_histogram(bins = 20) +
    geom_segment(data = crazy_abund, 
                 mapping = aes(x = n, y = rep(10, 3), 
                               xend = n, yend = rep(0, 3)), 
                 color = "red", 
                 arrow = arrow(length = unit(1, "char")),
                 inherit.aes = FALSE) +
    xlab("Maximum abundances") +
    ylab("Count") +
    this_ms_look()




```

<!-- Because these samples are so unusual we decided to drop them from all further analyses presented in the main text. However, we retain the outlier samples in the supplement to validate that our results are not biased by their removal. -->

```{r}
#| label: remove-outliers
#| dependson: [mete-calc]
#| eval: false

# remove from mete results
arth_mete_out <- arth_mete
arth_mete <- filter(arth_mete, 
                    !((site %in% outliers$site) & (tree %in% outliers$tree)))

# remove from arth data frame
arth_out <- arth
arth <- filter(arth, 
               !((site %in% outliers$site) & (tree %in% outliers$tree)))

```




## Deviations from METE across the chronosequence

Now we analyze how deviations from METE behave across the chronosequence. The following code produces the analysis of a hump-shaped versus flat relationship between METE deviation and flow age.

```{r}
#| label: z2-chrono-lm
#| dependson: [mete-calc, ]

# quadratic model of sad z^2 across age
sad_rma_chrono <- site_means_lm(log(z2_sad) ~ 
                                    log(flow_age_my) + I(log(flow_age_my)^2) + 
                                    (1 | site),
                                data = arth_mete)

# quadratic model of ipd z^2 across age
ipd_rma_chrono <- site_means_lm(log(z2_ipd) ~ 
                                    log(flow_age_my) + I(log(flow_age_my)^2) + 
                                    (1 | site),
                                data = arth_mete)

```

Likelihood ratio tests (LRT) can evaluate the significance of the quadratic model


```{r}
#| label: z2-chrono-lrt
#| dependson: [mete-calc, z2-chrono-lm]

# likelihood ratio tests for sad and ipd across chrono
sad_lrt_chrono <- anova(sad_rma_chrono)
sad_lrt_chrono

ipd_lrt_chrono <- anova(ipd_rma_chrono)
ipd_lrt_chrono
```

The quadratic model for SAD $z^2$-values is supported by the LRT with a $P$-value < 0.0001, but the quadratic model for IPD $z^2$-values did not have a significant $P$-value. However, the 95% CI for the quadratic term in the IPD $z^2$-values model has the majority of its range below zero, indicating the data do tend toward a quadratic shape, just with greater variance compared to the quadratic model for SAD $z^2$-values.

```{r}
#| dependson: [mete-calc, z2-chrono-lm]

ipd_rma_chrono

```


The following code produces the visualization reported in the main text of deviation from METE across flow ages.

```{r}
#| label: fig-z2-chrono
#| fig-height: 7
#| fig-cap: |
#|     Deviations from METE as measured by $z^2$ across the chronosequence. 
#|     Discussed in more detail in the main text.
#| dependson: mete-calc, z2-chrono-lm

# prepare data frame of predicted values for plotting
x <- range(arth$flow_age_my) |> 
    log() |> 
    (\(x) seq(x[1], x[2], length.out = 50))() |> 
    exp() |> 
    data.frame(flow_age_my = _)

sad_hat <- site_means_predict(sad_rma_chrono, newdata = x)
ipd_hat <- site_means_predict(ipd_rma_chrono, newdata = x)

z2_chrono_pred <- bind_rows(z2_sad = data.frame(x, sad_hat), 
                            z2_ipd = data.frame(x, ipd_hat), 
                            .id = "metric") |> 
    select(!se) |> 
    rename(z2 = pred,
           z2_lo = ci.lb, 
           z2_hi = ci.ub) |> 
    mutate(across(starts_with("z2"), exp))


# plotting
fig_z2_chrono <- pivot_longer(arth_mete, c(z2_sad, z2_ipd), 
             names_to = "metric", values_to = "z2") |> 
    ggplot(aes(flow_age_my, z2, color = site_name)) +
    geom_jitter(width = 0.1) +
    # confidence envelope 
    geom_ribbon(data = z2_chrono_pred, 
                mapping = aes(x = flow_age_my, 
                              ymin = z2_lo, 
                              ymax = z2_hi), 
                color = "transparent", alpha = 0.4) +
    # central tendency 
    geom_line(data = z2_chrono_pred, 
              mapping = aes(flow_age_my, z2), 
              color = "black") +
    # facet by IPD or SAD and trick strip label into axis label
    facet_grid(
        rows = vars(metric), 
        labeller = as_labeller(
            c(z2_ipd = "\"IPD\"~z^2*\"-value\"", 
              z2_sad = "\"SAD\"~z^2*\"-value\""),
            label_parsed
        ), 
        switch = "y"
    ) +
    xlab("Flow age (my)") +
    this_ms_look(log_x = TRUE, log_y = TRUE) +
    theme(axis.title.y = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside")

fig_z2_chrono

```




## How do observed patterns deviate from METE?

Next we determine how the observed data deviate from the METE predictions. To begin we look at two exemplar SADs and two exemplar IPDs representing low and high deviation from theory (@fig-how-diff).

```{r}
#| label: fig-how-diff
#| fig-cap: |
#|     Two exemplar IPDs and SADs showing high and low deviation from METE.
#|     The low deviation distributions come from the same sample at the 
#|     Alakaʻi site while the high deviation distributions come from the
#|     same sample at the Kamakou site. Samples were choosen to represent
#|     ranges of deviation while still being similar in state variables
#| dependson: [mete-calc]

# function to summarize theoretical and observed distributions
# given one sample `x`

mk_summ <- function(x) {
    esf <- meteESF(x$species_code, x$abundance, x$metabolic_rate)
    
    list(sad = sad(esf),
         ipd = ipd(esf)) |> 
        lapply(function(x) {
            thr <- samp_mete(x, n = 1000) |> 
                summarize_samp_mete()
            
            obs <- data.frame(rank = seq_along(x$data), 
                              abund_obs = x$data)
            
            full_join(thr, obs)
        }) |> 
        bind_rows(.id = "dist")
}



expl_samps <- data.frame(site = c("KA", "MO"), 
                         tree = c(5, 9), 
                         deviation = c("lo", "hi")) |> 
    left_join(arth) |> 
    (\(x) split(x, x$deviation, drop = TRUE))() |> 
    lapply(mk_summ) |> 
    bind_rows(.id = "deviation")


fig_how_diff <- ggplot(expl_samps, aes(rank, m)) +
    geom_point(aes(rank, abund_obs)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), 
                fill = "#489BC5", alpha = 0.75) +
    geom_line(color = "#489BC5") +
    facet_grid2(
        rows = vars(dist), cols = vars(deviation), 
        scales = "free", independent = "x", 
        labeller = as_labeller(
            c(hi = "High deviation", lo = "Low deviation", 
              ipd = "Metabolic rate", sad = "Abundance")
        ), 
        switch = "y"
    ) +
    xlab("Rank") +
    ylab("") +
    this_ms_look(log_y = TRUE) +
    theme(strip.background.y = element_blank(), 
          strip.placement.y = "outside",
          strip.text.y = element_text(size = 14))

fig_how_diff
```


Looking across all samples (@fig-how-z2) we see that deviations between observed and theoretical SADs is consistently in the direction of under-predicted both rarity and dominance (i.e. observed SADs have more rarity and greater dominance than theoretical predictions). Indeed, only one sample shows a (small) over-prediction in dominance, and only two samples show a (small) over-prediction in rarity. This means that observed SADs are more uneven than the METE prediction. 


```{r}
#| label: fig-how-z2
#| fig-width: 7.5
#| fig-height: 6
#| fig-cap: |
#|     Relationship between $z^2$ measure of deviation from METE and different
#|     metrics describing how observed data and METE predictions diverge.
#| dependson: [mete-calc]

fig_how_z2 <- bind_rows(sad = select(arth_mete, site_name, 
                       z2 = z2_sad, rar_diff = n1_diff, dom_diff = nmax_diff),
          ipd = select(arth_mete, site_name, 
                       z2 = z2_ipd, rar_diff = p1_diff, dom_diff = pmax_diff),
          .id = "dist") |> 
    pivot_longer(ends_with("diff"), 
                 names_to = c("comp", ".value"), 
                 names_sep = "_") |> 
    ggplot(aes(diff, z2, color = site_name)) +
    geom_point() +
    facet_grid2(rows = vars(dist), cols = vars(comp), 
                scales = "free", independent = "x", 
                labeller = as_labeller(
                    c(ipd = "\"IPD\"~z^2*\"-value\"", 
                      sad = "\"SAD\"~z^2*\"-value\"", 
                      dom = "\"Diff. in dominance\"", 
                      rar = "\"Diff. in rarity\""),
                    label_parsed
                ), 
                switch = "both") +
    xlab("") +
    ylab("") +
    this_ms_look(log_y = TRUE) +
    geom_vline(xintercept = 0, color = "gray") +
    theme(strip.background = element_blank(), 
          strip.placement = "outside",
          strip.text.y = element_text(size = 14))


fig_how_z2
```


In the case of the IPD dominance refers to metabolic rates of individuals with the highest metabolic rates, and rarity refers to the number of individuals with very low metabolic rates. Actual and theoretical IPDs do not show this consistent discrepancy between observed and predicted dominance and rarity. To evaluate this we model the probability of whether the log ratio is above or below zero with an intercept only binomial GLM. 

```{r}
#| label: ipd-how-glm
#| dependson: [mete-calc]

# "successes" and "failures" for GLM
ipd_diff <- summarize(arth_mete, 
                      n_over_dom = sum(pmax_diff < 0), 
                      n_undr_dom = sum(pmax_diff > 0), 
                      n_over_rar = sum(p1_diff < 0), 
                      n_undr_rar = sum(p1_diff > 0), 
                      .by = site)

# dominance 
pmax_mod <- glm(cbind(n_over_dom, n_undr_dom) ~ 1, 
                family = binomial, data = ipd_diff) 

summary(pmax_mod)


# rarity
p1_mod <- glm(cbind(n_over_rar, n_undr_rar) ~ 1, 
              family = binomial, data = ipd_diff)

summary(p1_mod)

```

The fact that in both models the intercept is not significantly different from the null ($p = 0.5$) means that there is not evidence of systematic difference between thertical prediction and observed IPD.


## Comparing deviations from METE with potential explanatory variables


### Deviation vs. proportion of non-native species

To see if proportion of non-native species has any relationship with deviation from METE we use meta regression as with the chronosequence analysis. The below code produces this meta regression and tests the affect of proportion of non-native species on deviations of the SAD and IPD from METE predictions.

```{r}
#| label: calc-dev-non
#| dependson: [mete-calc]

# calculate proportion of non-native spp per site
mete_by_non <- summarize(
    arth, 
    prop_nspp_non_nat = n_distinct(
        species_code[!(origin %in% c("endemic", "indigneous"))]) / 
        n_distinct(species_code),
    .by = site_name) |> 
    # join to mete summary statistics
    full_join(arth_mete)

# sad regression 
sad_rma_non <- site_means_lm(log(z2_sad) ~ prop_nspp_non_nat + (1 | site), 
                             # excluding Kīlauea site because it is 
                             # quite different, further assessed below
                             filter(mete_by_non, site != "VO"))

# sad likelihood ratio test
anova(sad_rma_non)


# ipd regression 
ipd_rma_non <- site_means_lm(log(z2_ipd) ~ prop_nspp_non_nat + (1 | site), 
                             # excluding Kīlauea site
                             filter(mete_by_non, site != "VO"))

# ipd likelihood ratio test
anova(ipd_rma_non)

```

The below code is used for visualizing the results. Note that Kīlauea is considered separately because it falls clearly below the expected trend based on other sites. In the case of the SAD, the mean and 95% CI of the $z^2$-value for Kīlauea is below both the expected value and 95% CI of the rest of the data. In the case of the Kīlauea IPD only the mean of $z^2$-values is below the 95% CI of the rest of the data; the two CIs do overlap.  This is why in the meta regression analysis Kīlauea was excluded, so we could interrogate the way it deviates from the overall pattern.

```{r}
#| label: fig-dev-non
#| fig-height: 7
#| fig-cap: |
#|     Relationship between proportion of non-native species and deviation from 
#|     METE. Kīlauea was excluded from the overall analysis to be able to 
#|     compare it separately to the overall trend. Gray regreions represent
#|     95% confidence envelopes (or confidence interval for Kīlauea) and solid 
#|     lines are expected values.
#| dependson: [mete-calc, calc-dev-non]


# 95% CI for mean of Kīlauea site
vo_mean <- filter(mete_by_non, site == "VO") |> 
    pivot_longer(starts_with("z2"), 
                 names_to = c(".value", "dist"), names_sep = "_") |> 
    mutate(z2 = log(z2)) |> 
    summarize(site_name = first(site_name), 
              prop_nspp_non_nat = mean(prop_nspp_non_nat),
              z2_mean = mean(z2), 
              z2_se = sd(z2) / sqrt(n()), 
              z2_lo = z2_mean - 1.96 * z2_se,
              z2_hi = z2_mean + 1.96 * z2_se, 
              .by = dist) |> 
    rename(z2 = z2_mean) |> 
    mutate(across(starts_with("z2"), exp))

# prediction for plotting
x <- range(mete_by_non$prop_nspp_non_nat) |> 
    (\(x) seq(x[1], x[2], length.out = 50))() |> 
    data.frame(prop_nspp_non_nat = _)

sad_hat <- site_means_predict(sad_rma_non, newdata = x)
ipd_hat <- site_means_predict(ipd_rma_non, newdata = x)

z2_non_pred <- bind_rows(sad = data.frame(x, sad_hat), 
                         ipd = data.frame(x, ipd_hat), 
                         .id = "dist") |> 
    select(!se) |> 
    rename(z2 = pred,
           z2_lo = ci.lb, 
           z2_hi = ci.ub) |> 
    mutate(across(starts_with("z2"), exp))

# plotting
fig_dev_non <- pivot_longer(mete_by_non, starts_with("z2"), 
                            names_to = c(".value", "dist"), 
                            names_sep = "_") |> 
    ggplot(aes(prop_nspp_non_nat, z2, color = site_name)) +
    geom_jitter(width = 0.002) +
    # confidence envelope
    geom_ribbon(data = z2_non_pred, 
                mapping = aes(x = prop_nspp_non_nat, 
                              ymin = z2_lo, 
                              ymax = z2_hi), 
                color = "transparent", alpha = 0.4) +
    # central tendency 
    geom_line(data = z2_non_pred, 
              mapping = aes(prop_nspp_non_nat, z2), 
              color = "black") +
    geom_rect(data = vo_mean, 
              mapping = aes(xmin = prop_nspp_non_nat - 0.002, 
                            xmax = prop_nspp_non_nat + 0.002, 
                            ymin = z2_lo, 
                            ymax = z2_hi), 
              color = "transparent", alpha = 0.4) +
    geom_segment(data = vo_mean, 
              mapping = aes(x = prop_nspp_non_nat - 0.002, 
                            xend = prop_nspp_non_nat + 0.002, 
                            y = z2, 
                            yend = z2), 
              color = "black") +
    facet_grid(rows = vars(dist), 
               labeller = as_labeller(
                   c(ipd = "\"IPD\"~z^2*\"-value\"", 
                     sad = "\"SAD\"~z^2*\"-value\""),
                   label_parsed
               ), 
               switch = "y") +
    xlab("Proportion of non-native spp.") +
    this_ms_look(log_y = TRUE) +
    theme(axis.title.y = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside")



fig_dev_non
```



### Deviation vs. dissimilarity

We also conducted meta regression to see if $z^2$-values change meaningfully across different levels of within-site, between-sample dissimilarity or $\beta$-diversity. The below code runs this meta regression analysis.

```{r}
#| label: calc-diss-z
#| dependson: [mete-calc]

# calculate dissimilarity z-scores
diss <- split(arth, arth$site) |> 
    sapply(FUN = function(dat) {
        summarize(dat, abund = n(), .by = c(tree, species_code)) |> 
            pivot_wider(names_from = species_code, 
                        values_from = abund, 
                        values_fill = 0) |> 
            select(!tree) |> 
            bray_diss_z() 
    })

# combine with mete summary statistics
mete_by_diss <- mutate(arth_mete, 
                       mean_diss = diss[site])


# sad regression 
sad_rma_diss <- site_means_lm(log(z2_sad) ~ mean_diss + (1 | site),
                              data = mete_by_diss)

# sad likelihood ratio test
anova(sad_rma_diss)


# ipd regression 
ipd_rma_diss <- site_means_lm(log(z2_ipd) ~ mean_diss + (1 | site),
                              data = mete_by_diss)

# ipd likelihood ratio test
anova(ipd_rma_diss)


```

The relationship between deviation from METE and dissimilarity is more clearly linear than that of proportion non-native species and deviation from METE. It is therefore worth figuring out if the ways that native and non-native taxa contribute to dissimilarity is different across sites.


The below code visualizes the results of the meta regression or SAD and IPD $z^2$-values against dissimilarity.

```{r}
#| label: fig-diss-z
#| fig-height: 8
#| fig-cap: |
#|     Relationship between dissimilarity and deviation from METE. Gray 
#|     regreions represent 95% confidence envelopes and solid lines are 
#|     expected values.
#| dependson: [mete-calc, calc-diss-z]

# prediction for plotting
x <- range(mete_by_diss$mean_diss) |> 
    (\(x) seq(x[1], x[2], length.out = 50))() |> 
    data.frame(mean_diss = _)

sad_hat <- site_means_predict(sad_rma_diss, newdata = x)
ipd_hat <- site_means_predict(ipd_rma_diss, newdata = x)

z2_diss_pred <- bind_rows(sad = data.frame(x, sad_hat), 
                         ipd = data.frame(x, ipd_hat), 
                         .id = "dist") |> 
    select(!se) |> 
    rename(z2 = pred,
           z2_lo = ci.lb, 
           z2_hi = ci.ub) |> 
    mutate(across(starts_with("z2"), exp))

# plotting
fig_diss_z <- pivot_longer(mete_by_diss, starts_with("z2"), 
                            names_to = c(".value", "dist"), names_sep = "_") |> 
    ggplot(aes(mean_diss, z2, color = site_name)) +
    geom_jitter(width = 0.1) +
    # confidence envelope
    geom_ribbon(data = z2_diss_pred, 
                mapping = aes(x = mean_diss, 
                              ymin = z2_lo, 
                              ymax = z2_hi), 
                color = "transparent", alpha = 0.4) +
    # central tendency 
    geom_line(data = z2_diss_pred, 
              mapping = aes(mean_diss, z2), 
              color = "black") +
    facet_grid(rows = vars(dist), 
               labeller = as_labeller(
                   c(ipd = "\"IPD\"~z^2*\"-value\"", 
                     sad = "\"SAD\"~z^2*\"-value\""),
                   label_parsed
               ), 
               switch = "y") +
    xlab("Mean dissimilarity") +
    this_ms_look(log_y = TRUE) +
    theme(axis.title.y = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside")

fig_diss_z


```


### Partitioning Bray Curtis dissimilarity between native and non-native species

To do that we take advantage of the additive nature of Bray Curtis dissimilarity. Bray Curtis dissimilarity is defined as

$$
d_{kj} = \frac{\sum_i |x_{ij} - x_{ik}|}{\sum_{ij} x_{ij} + x_{ik}}
$$

where $x_{ij}$ and $x_{ik}$ are the abundances of species $i$ at sites $j$ and $k$. This can be partitioned across any grouping of species, say $(a)$ and $(b)$ with abundances $x_{ij}^{(a)}$ and $x_{ij}^{(b)}$

$$
d_{kj} = \frac{\sum_i |x_{ij}^{(a)} - x_{ik}^{(a)}|}{\sum_{ij} x_{ij} + x_{ik}} + \frac{\sum_i |x_{ij}^{(b)} - x_{ik}^{(b)}|}{\sum_{ij} x_{ij} + x_{ik}}
$$

Now we can look separately at the contribution of group $(a)$ and group $(b)$ to the overall dissimilarity. We use this approach to compare the contributions of native and non-native taxa to overall dissimilarity. We again make use of a permutational null model to compare these contributions. This null model cannot be the same as the null model used to calculate z scores for overall dissimilarity because that null model shuffled species across samples and individuals across species. Instead, we want to shuffle group labels (native or non-native) across species. The custom function `bray_part_z` calculates these z scores using another custom function `cswap` from *HawaiiArthMETE*.

```{r}
#| label: bray-part
#| dependson: [mete-calc]

site_bray_z <- split(arth, arth$site_name) |> 
    lapply(function(dat) {
        comm <- summarize(dat, 
                  abund = n(), 
                  .by = c(tree, species_code)) |> 
            pivot_wider(names_from = species_code, 
                        values_from = abund, 
                        values_fill = 0) |> 
            select(!tree)
        g <- dat$origin[match(names(comm), dat$species_code)] %in% 
            c("endemic", "indigenous")
        
        data.frame(bp = bray_part_z(as.matrix(comm), g))
    }) |> 
    bind_rows(.id = "site_name")


```


Now we can visualize different contributions of native and non-native species to within-site, between-sample dissimilarity.

```{r}
#| label: fig-bray-part
#| fig-cap: |
#|     Difference in contribution of native and non-native species to 
#|     dissimilarity compared to null. Negative values indicate greater 
#|     contribution of non-native species compared to chance while positive
#|     values indicate greater contribution of native species compared to 
#|     chance.
#| dependson: [mete-calc, bray-part]

site_bray_z$site_name <- factor(site_bray_z$site_name, 
                                levels = levels(arth$site_name))

fig_bray_part <- ggplot(site_bray_z, aes(bp, site_name, fill = site_name)) +
    geom_bar(stat = "identity") +
    this_ms_look() +
    geom_vline(xintercept = 0, color = "gray") +
    xlim(c(-1, 1) * max(abs(site_bray_z$bp))) +
    ylab("") +
    xlab("Log ratio dissimilarity partition") +
    annotate("segment", x = c(-0.5, 0.5), xend = c(-2.75, 2.75), 
             y = 6, yend = 6, 
             arrow = arrow(length = unit(1, "char"))) +
    annotate("text", x = c(-3, 3), y = 6, 
             label = c("More non-native\ncontribution", 
                       "More native\ncontribution"), 
             hjust = c(1, 0), size = 4) +
    theme(legend.position = "none", 
          plot.margin = margin(40, 40, 7, 7)) +
    coord_cartesian(ylim = c(1, 5), clip = "off") 

fig_bray_part
```



