---
title: "Eco-evolutionary deviations from steady state across the Hawaiian chronosequence revealed with a maximum entropy perspective"
author:
  - name: Andrew J. Rominger
    orcid: 0000-0003-3755-4480
    affiliations:
      - name: University of Hawaiʻi at Mānoa
  - name: Katie Thai
    orcid: 0009-0006-1223-122X
    affiliations:
      - name: University of Hawaiʻi at Mānoa
  - name: Rosemary G. Gillespie
    orcid: 0000-0003-0086-7424
    affiliations:
      - name: University of California Berkeley
  - name: Daniel S. Gruner
    orcid: 0000-0002-3153-4297
    affiliations:
      - name: University of Maryland
  - name: John Harte
    orcid: 0000-0002-8794-5140
    affiliations:
      - name: University of California Berkeley
filters:
  - authors.lua
format:
  pdf:
    keep-tex: true
    include-in-header: 
      text: |
        \usepackage{authblk}
        \usepackage{orcidlink}
        \usepackage{newunicodechar}
        \newunicodechar{ʻ}{`} 
        \usepackage{xr}
        \externaldocument{hawaii-arth-mete-supp}
        \newcommand{\sfig}[1]{Supplementary Figure~\ref{#1}}
        \newcommand{\ssec}[1]{Supplementary Section~\ref{#1}}
        \renewcommand{\sfdefault}{\rmdefault}
knitr:
  opts_chunk: 
    dev: cairo_pdf
csl: ecology-letters.csl
bibliography: hawaii-arth-mete.bib
---


```{r}
#| label: setup
#| include: false


library(ggplot2)
library(cowplot)
library(grid)
library(sf)
library(dplyr)
library(metafor)
library(meteR)
library(HawaiiArthMETE)

knitr::opts_chunk$set(fig.width = 5.5, fig.height = 4, 
                      echo = FALSE, message = FALSE, warning = FALSE)

```

## Introduction

Biodiversity patterns are bound by two seemingly contradictory states: they show striking consistency in form across systems, and yet also must emerge in spite of, or because of, the constant flux and dynamic change of ecology and evolution. The near universality of biodiversity patterns includes the scaling of species richness with area, the ubiquity of rare spaces and scarcity of common species, and the highly skewed distribution of biomass or metabolic resource use across organisms. Debate has persisted for over a century as to whether these consistent patterns form because of ecological and evolutionary drivers that force systems into these configurations, or are largely the outcome of chance acting on large systems of constituent parts. The former gives rise to niche-based theories of coexistence, while that later has produced a number of neutral-like models.

Both deterministic and stochastic perspectives on the generation of universal biodiversity patterns most often make the choice of considering an ecosystem at "equilibrium" or steady state. However, ecosystems are known to be dynamic both through intrinsic eco-evolutionary processes and due to recent anthropogenic perturbation. Whether universal patterns are insensitive to non-steady-state processes or are a result of those processes remains addressed. To better understand the interaction of non-steady-state processes and universal patterns, we combine data from the highly dynamic ecosystems of the Hawaiian Archipelago with the intentionally minimalist steady state predictions of the Maximum Entropy Theory of Ecology.

### The Hawaiian Chronosequence 

The Hawaiian Archipelago holds numerous examples of adaptive radiation [@gillespie2004; @knope2012; @baldwin1998; @magnacca2015; @givnish2009; @lerner2011; @bennett2013]. Those remarkable examples of diversification along with the geographic isolation and chronological geology of the archipelago have made it a focal point for the study of eco-evolutionary processes [@simon1987; @wagner-funk1995; @chadwick1999; @price2004; @gillespie_clague2009; @barton2021]. The manner in which the islands formed in chronological sequence [@fig-chrono] has allowed for the study of ecosystems in different stages of eco-evolutionary assembly [@vitusek2004; @gillespie2004; @rominger2016] by considering ecosystems on different aged substrates to be snapshots of eco-evolutionary processes at different time points.


```{r}
#| label: fig-chrono
#| results: hide
#| cache: true
#| fig-cap: |
#|     Chronosequence of the Hawaiian Archipelago showing oldest flows to the
#|     west on Kauaʻi and youngest flows to the east on Hawaiʻi. Substrate age
#|     data from @sherrod2021. The five sites are indicated with dots on the map and 
#|     arrows indicating their flow ages on the legend. From oldest to youngest 
#|     (west to east) sites are: Alakaʻi, Kamakou, Kohala, Laupāhoehoe, Kīlauea.

x <- st_read("../hi_geol/hawaii_state_geol_ageClean.shp") |> 
    filter(!is.na(age_mid)) 
hi <- st_read("../hi_geol/hi_outline.shp")
sites <- st_read("../hi_geol/gruner_sites.kml") |> 
    st_transform(crs = st_crs(x))

p <- ggplot(hi) +
    geom_sf(color = "gray30", fill = "gray30", linewidth = 0.1) +
    geom_sf(data = x,
            aes(fill = age_mid, color = age_mid),
            linewidth = 0.0001,
            inherit.aes = FALSE) +
    scale_fill_viridis_c(option = "magma", begin = 0.2, end = 1,
                         transform = c("log10", "reverse")) +
    scale_color_viridis_c(option = "magma", begin = 0.2, end = 1,
                         transform = c("log10", "reverse")) +
    geom_sf(data = sites, shape = 21, fill = "black", color = "white") +
    theme_cowplot() + 
    theme(legend.position = "none", 
          axis.line = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())


log_age_range <- range(log(x$age_mid, 10))
log_age <- pretty(log_age_range) |> 
    (\(x) x[x > log_age_range[1] & x < log_age_range[2]])()


l <- data.frame(y = seq(log_age_range[1], 
                   log_age_range[2], 
                   length.out = 500), 
           z = 1) |> 
    mutate(u = 10^y) |>
    ggplot(aes(z, y, fill = u, color = u)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", begin = 0.2, end = 1, 
                         transform = c("log10", "reverse")) +
    scale_color_viridis_c(option = "magma", begin = 0.2, end = 1, 
                         transform = c("log10", "reverse")) +
    annotate("segment",
        x = rep(c(0.47, 1.53), each = length(log_age)),
        xend = rep(c(0.7, 1.3), each = length(log_age)),
        y = rep(log_age, 2),
        yend = rep(log_age, 2), 
        color = "white", linewidth = 0.5) +
    annotate("segment",
        x = rep(-0.1, length(unique(arth$flow_age_my))),
        xend = rep(0.49, length(unique(arth$flow_age_my))),
        y = log(unique(arth$flow_age_my), 10),
        yend = log(unique(arth$flow_age_my), 10), 
        arrow = arrow(length = unit(0.25, "npc"))) +
    annotate("text", 
             x = 0.75, y = 1.5, 
             label = "Age (mya)") +
    scale_x_continuous(expand = FALSE) +
    scale_y_continuous(
        expand = FALSE,
        limits = c(-10, 5),
        sec.axis = sec_axis(
            ~ ., 
            breaks = rev(log_age),
            labels = sapply(rev(log_age), function(a) bquote(10^.(a)))
        )
    ) +
    theme_cowplot() +
    theme(legend.position = "none",
          axis.ticks = element_blank(), 
          axis.line = element_blank(), 
          axis.title.y = element_blank(), 
          axis.title.x.bottom = element_blank(),
          axis.text.x = element_blank(), 
          axis.text.y.left = element_blank(), 
          axis.text.y.right = element_text(size = 10)) + 
    coord_cartesian(clip = "off")


plot_grid(p, l, rel_widths = c(1, 0.2))
```

Hawaiʻi has also been subjected to pervasive anthropogenic changes to its biodiversity, especially a rapidly expanded human land-use footprint and markedly accelerated introduction of non-native species following Western contact and settler colonialism [@gon2018; @graham2017]. As such Hawaiʻi has produced numerous studies of rare and endangered species [@sakai2002; @magnacca2007; @pratt2009; @regnieretal2015] and the functions of non-native species in ecosystems [@ostertag2009; @barton2021].

Arthropods in Hawaiʻi represent both many adaptive radiations [@gillespie2004; @magnacca2015; @bennett2013], many rare species [@magnacca2007; @magnacca2015], and many introduced species [@hichecklist]. Given their abundance and diversity, they also present a unique opportunity to use ecological theory to study how biodiversity conforms to or deviates from steady state.


### The Maximum Entropy Theory of Ecology

The Maximum Entropy Theory of Ecology (METE @harte2011) is a particularly useful tool for studying steady state biodiversity patterns because it produces multiple testable predictions, making falsifications of the theory more informative [@mcgill2003] including the distribution of abundances across species and the distribution of metabolic rates across individuals in a sample. METE arrives at some similar predictions to other null or neutral models of biodiversity such at the Unified Neutral Theory [@hubbell2001] but instead of being limited to a single neutral formulation, METE is grounded in the principles of statistical mechanics [@harte2011 @jaynes1957] drawing from the probabilistic properties of large, randomly assembled systems. Thus its predictions represent a community in statistical steady state. Statistical steady state means that a system is governed by only a few simple state variables, which constitute a state space, and that no additional processes limit the system's ability to freely explore this state space. The state variables of METE are total number of species $S_0$, total number of individuals $N_0$, and summed metabolic rate across all individuals $E_0$ [@harte2011]. METE posits that if state variables are in steady state, all biodiversity patterns are statistical outcomes from those state variables independently of the specific biological mechanisms governing the values of the state variables, so long as those mechanisms do not constitute information beyond the state variables themselves. Examples of such additional information could be rapid environmental change leading to mismatched niches with local environments, rapid diversification, or biological invasion. Under such scenarios the simple state variables of $S_0$, $N_0$, and $E_0$ would not likely constitute enough information to yield accurate predictions for biodiversity patterns such as the species abundance distribution or distribution of metabolic rates across individuals. Thus finding deviations between observed patterns of abundance and metabolic rate distributions compared to their METE predictions allows us to identify ecological systems out of statistical steady state [@harte2011; @rominger2016; @newman2020]. 

In this study of arthropods communities across the Hawaiian chronosequence, we find consistent deviations from steady state at middle aged sites. Deviations are predicted by community dissimilarity which peaks at middle aged sites. This suggests these communities are unstable, perhaps due to their age being the right time for adaptive radiation to take off for arthropods [@gillespie2004; @bennett2013]. Deviations are also partially predicted by proportion of non-native species, except that the proportion of non-native species is high at the youngest site while this site also shows some of the lowest deviations from METE. This suggests the role of biological invasion is context dependent, at times producing stead state and at others driving deviation from it. More work connecting deviations from steady state to ecological and evolutionary drivers will be needed to further test these possibilities.


## Methods

### Reproducibility and data preparation

A custom R package, *HawaiiArthMETE* [@HawaiiArthMETE], accompanies this paper and can be used to reproduce all results. All data analyzed here are available via the package as described in \ssec{sec-setup}. 

Arthropod data come from the original publication by @gruner2007, a study that collected `r nrow(arth)` specimens across the Hawaiian chronosequence [@fig-chrono]. Study sites in @gruner2007 where standardized to be climatically homogeneous, seeking to capture differences in arthropod assemblages due to substrate age and not other abiotic factors. 

Each specimen was identified to the species level (morpho species were used in `r round(100 * mean(grepl("sp_", arth$species)), 2)`% of specimens). Specimen lengths were measured to the nearest millimeter.  For METE analyses we need to estimate metabolic rate from these data. Because METE normalizes metabolic rates such that the lowest possible rate is 1 we use a simplified $3/4$ scaling rule to approximate metabolic rate $B$ from mass $M$ as $B \propto M^{3/4}$.

This leaves the estimation of mass from length. @gruner2007 used taxon- and stage- specific length-to-biomass allometric scaling relationships [@gruner2003] to approximate mass as the expected value of a log-log regression

$$
\begin{aligned}
\text{mass} & = e^{\widehat{\log(\text{mass})}} \\
\widehat{\log(\text{mass})} &= \log(a) + b \cdot \log(\text{length})
\end{aligned}
$$

where $a$ and $b$ are estimated from data collected by @gruner2003 and vary by taxon and stage.

Using these expected value masses based on length measured to the nearest milliliter yields metabolic rate approximations that are artificially discretization (see the appendix "Sampling biomass estimates from allometric equations"). These descretizations in turn lead to spurious conclusions about the predicitiveness of METE because METE assumes metabolic rates are continuous (see \ssec{sec-re-run}).

Therefore, instead of estimating mass as the expected value from an allometric relationship we sample masses as random draws from the full model relating mass to length

$$
\begin{aligned}
\text{mass} & = e^{y} \\
y &\sim \mathscr{N}(\widehat{\log(\text{mass})}, ~\sigma) \\
\widehat{\log(\text{mass})} &= \log(a) + b \cdot \log(\text{length})
\end{aligned}
$$

where $\sigma$ is the taxon- and stage-specific standard deviation of residuals. The full procedure for sampling masses is detailed in the appendix "Sampling biomass estimates from allometric equations" and the consequences of using random draws versus expected values on the METE analysis is thoroughly investigated in \ssec{sec-re-run}. We conclude that using random draws for mass estimates is largely similar to using expected values, and our overall conclusions are unchanged, but using draws avoids both spurious deviations from and conformations to METE that would be introduced by using expected values of mass.

### METE analysis

To produce METE predictions for the SAD and IPD we use the *meteR* R package [@rominger-merow2016]. We compare these predictions to observed SAD and IPD patterns using a $z^2$-value. The use of $z^2$-values is discussed in @rominger-merow2016, but briefly: the $z^2$-value compares the log likelihood of the data given them METE model to the log likelihood of a hypothetical data set drawn from the METE model itself. This comparison is done many times and the actual $z^2$-value is calculated as

$$
z^2 = \left(\frac{\log(\mathscr{L}(\text{data} \mid \text{METE})) - \hat{\mu}_{\log\mathscr{L}_\text{METE}}}{\hat{\sigma}_{\log\mathscr{L}_\text{METE}}}\right)^2
$$

where $\hat{\mu}_{\log\mathscr{L}_\text{METE}}$ is the mean of the log likelihood of data drawn from METE itself, and $\hat{\sigma}_{\log\mathscr{L}_\text{METE}}$ is the standard deviation of the log likelihood of data drawn from METE itself.

If METE is truly correct in describing the data, $z^2$ will asymptotically have a $\chi^2$ distribution with $d.f. = 1$ [@rominger-merow2016]. If METE does not truly describe the data, the $z^2$-value is likely to be large.

We also quantify the difference between METE predictions and observed SADs and IPDs by comparing the tails of the observed and predicted distributions. These tails can be especially informative about the nature of deviations [@harte2011]. For the SAD we compare rarity as the log ratio of observed proportion of singletons to the predicted proportion of singletons, and we compare dominance as log ration of observed total abundance of the top three most abundant species to the predicted total abundance of the top three most abundant species. For the IPD we compare metrics similar to rarity and dominance: the log ratio of observed proportion of individuals with a scaled metabolism of less than 3 to its prediction, and the log ratio of the observed sum of metabolic rates for the top 10 individuals to its prediction.



## Statistical analysis of deviations from METE and potential explanatory variables

We first evaluate how deviations from METE, as measured by $z^2$-values, vary across the chronosequence. We then test for relationships between $z^2$-values and proportion of non-native species at a site as well as within site, between sample dissimilarity or $\beta$-diversity.

For all these analyses we fit linear models predicting log-transformed $z^2$-values responding to each explanatory variable: substrate age, proportion non-native species, and within site $\beta$-diversity. Log-transforming $z^2$-values brings them closer to the normal assumptions of a linear model. The need for log-transformation is not surprising: while $z^2$-values will be $\chi^2$ distributed if METE is true, deviations from METE add more probability to larger values, and taken together will lead to right-skewed distribution of $z^2$-values. Log-transformation brings right-skewed random variables closer to normal.

When modeling deviations from METE, we have two statistical challenges to address: samples within sites could represent pseudo replicates necessitating a random-effects model; however, because explanatory variables are site-level characteristics, site-level random effects would be completely collinear with site-level explanatory variables. 

To address these challenges we model the site-level mean and variance rather than each individual sample within each site. This is the most conservative approach but runs the risk of being under-powered because while we have `r nrow(unique(arth[, c("site", "tree")]))` samples, we only have `r n_distinct(arth$site)` sites. Modeling mean and variance is the same technique used in meta-analyses. We therefore make use of the *metafor* package [@metafor] in fitting testing these models.

## Calculating $\beta$-diversity

To quantify within site, between sample $\beta$-diversity we first calculate mean Bray-Curtis dissimilarity between samples and then, to standardize $\beta$-diversity measures across sites, we normalize mean dissimilarity to a null model. We use the `quasiswap_count` algorithm implemented in *vegan* [@vegan] to generate null samples. This algorithm preserves row and column sums of a sample by species matrix and also seeks to maintain null matrix fill close to the observed fill thereby mimicking patchiness in the real data. We use 500 null sample by species matrices to generate null models for each site. Then normalized within site, between sample $\beta$-diversity was calculated as a z-score

$$
\text{Normalized }\beta = \frac{\text{BC}_\text{obs} - \mu_\text{null}}{\sigma_\text{null}}
$$

where $BC_\text{obs}$ is the observed mean Bray-Curtis dissimilarity between samples within a site, $\mu_\text{null}$ is the mean of the null distribution, and $\sigma_\text{null}$ is the standard deviation of the null distribution.

We are also interested in understanding how native and non-native species contribute to the overall mean dissimilarity at a site. We can do this by partitioning Bray Curtis dissimilarity between native and non-native groups of species. \ssec{sec-part} provides the mathematical and computational details of how we do this. The end result of partitioning Bray-Curtis dissimilarity between native and non-native species is that we have a ratio of within site $\beta$-diversity attributable to native species to that attributable to non-native species. We take the log of this ratio such that equal contributions of native and non-native species yields a log ratio of 0.

We also compare this Bray Curtis dissimilarity partitioning to a null model. Here the `quasiswap_count` algorithm will not work because it permutes individuals across species as well as sites. Instead we want to only shuffle the label of each species: native or non-native. But we still need to maintain, at least probabilistically, attributes of the data that might confound null model analysis. Specifically, we should seek to maintain, in probability, the distribution of abundances of native and non-native species. We wrote a custom algorithm to do this in the *HawaiiArthMETE* package and described in \ssec{sec-bart}.

## Results

```{r}
#| label: supp-read
#| include: false

ff <- list.files("hawaii-arth-mete-supp_cache/pdf/", pattern = ".RData", 
                 full.names = TRUE) |> 
    gsub(".RData", "", x = _)

for(f in ff) {
    lazyLoad(f)
}
```

### Deviations from METE across the chronosequence

```{r}
#| label: fig-z2-chrono
#| fig-cap: |
#|     Deviations from METE across the chronosequence as measured by 
#|     $z^2$-values. Colors represent sites and match the color coding of 
#|     substrate age in @fig-chrono.

fig_z2_chrono

```

Deviations from METE in the SAD an IPD are largely in agreement: samples with high $z^2$-values for the SAD tend to have high $z^2$-values for the IPD (see \sfig{fig-z2-cor}). The $z^2$-values for both distributions peak at middle aged sites along the chronosequence [@fig-z2-chrono]. For the SAD this is supported by a likelihood ratio test comparing a quadratic model of $z^2$-values to an intercept-only model with LRT statistic $\Lambda =$ `r sad_lrt_chrono$QM`, $d.f. =$ `r sad_lrt_chrono$QMdf[1]`, and $P$ `r ifelse(sad_lrt_chrono$QMp < 0.0001, "< 0.0001", paste("= ", round(sad_lrt_chrono$QMp, 4)))`. 

The overall spread of IPD $z^2$-values is greater than that of SAD $z^2$-values. Not surprisingly, for quadratic model of IPD $z^2$-values across the chronosequence, the LRT statistic is $\Lambda =$ `r ipd_lrt_chrono$QM` with $d.f. =$ `r ipd_lrt_chrono$QMdf[1]` and $P$ `r ifelse(ipd_lrt_chrono$QMp < 0.0001, "< 0.0001", paste("= ", round(ipd_lrt_chrono$QMp, 4)))`. While the $P$-value for the test of the quadratic model for IPD $z^2$-values just misses the $\leq 0.05$ significance cut off by the likelihood ratio test, the less conservative Wald test does show the quadratic term to be significant with a 95% confidence interval of `r confint(ipd_rma_chrono, fixed = TRUE, random = FALSE) |> (`[[`)("fixed") |> (`[`)(3, 2:3) |> round(3) |> paste(collapse = ", ") |> (\(x) sprintf("[%s]", x))()`, which does not overlap 0.

For the SAD, the primary driver of differences between METE predictions and observed patterns is that observed species abundances have both a higher proportion of singletons and a larger concentration of abundance in the most abundant species [@fig-how-z2] compared to METE. Taken together, this means that observed SADs are more uneven than those predicted by METE.

For the IPD there is not a consistent relationship between deviations and the tails of the metabolic rate distribution [@fig-how-z2]. In terms of proportion of individuals with very low metabolisms, `r sum(ipd_diff[, "n_over_rar"])` samples showed theory over-predicting this proportion while `r sum(ipd_diff[, "n_undr_rar"])` samples showed theory under-predicting this proportion. The difference was not significant by a binomial generalized linear model of over-versus-under prediction grouped by site (z = `r round(summary(p1_mod)$coefficients[3], 3)`, $P$ = `r `round(summary(p1_mod)$coefficients[4], 3)`). In terms of concentration of metabolic rate in the most metabolically active individuals, `r sum(ipd_diff[, "n_over_dom"])` samples showed theory over-predicting this concentration while `r sum(ipd_diff[, "n_undr_dom"])` samples showed theory under-predicting this concentration. With the two numbers exactly equal, the difference could not be significant.


```{r}
#| label: fig-how-z2
#| fig-width: 7
#| fig-height: 5
#| fig-cap: |
#|     Relationship between $z^2$ measure of deviation from METE and different
#|     metrics describing how observed data and METE predictions diverge.

fig_how_z2
```


## Deviations from METE predicted by non-native species and $\beta$-diversity


```{r}
#| label: fig-pred
#| fig-width: 7
#| fig-height: 5
#| fig-cap: |
#|     Relationship between $z^2$ measure of deviation from METE and different
#|     predictor variables: proportion of non-native species and 
#|     $\beta$-diversity.

f_non <- fig_dev_non +
    theme(legend.position = "none", 
          plot.margin = unit(c(7, 1, 7, 7), "pt"))

f_dis <- fig_diss_z +
    theme(axis.text.y = element_blank(), 
          axis.ticks.length.y = unit(0, "npc"), 
          strip.text = element_blank(), 
          plot.margin = unit(c(7, 7, 7, 1), "pt"))

plot_grid(f_non, f_dis, rel_widths = c(1, 1.25))

```

- deviations are predicted by proportion of non-natives
- deviations are predicted by dissimilarity

## Partitioning $\beta$-diversity across native and non-native species


```{r}
#| label: fig-bray-part
#| fig-cap: |
#|     Difference in contribution of native and non-native species to 
#|     dissimilarity compared to null. Negative values indicate greater 
#|     contribution of non-native species compared to chance while positive
#|     values indicate greater contribution of native species compared to 
#|     chance.

fig_bray_part
```

- something is going on at Kīlauea


## Discussion

- dissimilarity as instability
- possible causes of instability---adaptive radiation
- presents future hypothesis to test with genetic data
- hypotheses
    - diversification drives divergence; diversification slowed down on older flows
    - invasion drives divergence
    - both do and they interact, perhaps even conditions favoring diversification favor invasion too
    
## References
