---
title: "Analysis of arthopod communites in Hawaii using METE Theory"
format: 
    pdf:
        include-in-header:
            text: |
                \usepackage{newunicodechar}
                \newunicodechar{ʻ}{`} 
---


This supplement walks through the steps of recreating the entire analysis presented in the main text. It also provided further analysis in support of and to contextualize results in the main text. 

## Computing setup

Analyses are supported by a custom R package *HawaiiArthMETE* [@foo] which can be installed from github. 

```{r}
#| label: pkg-install
#| eval: false

devtools::install_github("ecoevomatics/HawaiiArthMETE")
```

The quarto [@quarto] document used to generate this supplement can be found with the R command `system.file("ms/hawaii-arth-mete-supp.qmd", package = "HawaiiArthMETE")`.

The following R computing environment is used for all analyses:

```{r}
#| label: setup

library(meteR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vegan)
library(metafor)
library(HawaiiArthMETE)


knitr::opts_chunk$set(
    # default figure dims
    fig.width = 5.5, fig.height = 4, 
    # caching is default for all subsequent chunks
    cache = TRUE
)

# load data 
data(arth)

sessionInfo() |> 
    print(locale = FALSE)
```


## METE calculations

Comparison of the observed data to predictions from the Maximum Entropy Theory of Ecology (METE) is performed with package *meteR* [@rominger-merow2016]. The following code records information about each sample (name, flow age, proportion non-native presence) as well as several metrics of deviation between theory and observation:

- a $z^2$ value measuring a normalized difference in likelihoods between observed data and theoretical expectation
- differences in proportion of rarity observed and expected
- differences in proportion of dominance observed and expected

All metrics are calculated for both the species abundance distribution (SAD) and individual metabolic rate (or power) distribution (IPD).

```{r}
#| label: mete-calc
#| message: false
#| output: false
#| cache: true

arth_mete <- split(arth, arth[, c("site", "tree")], drop = TRUE) |> 
    lapply(function(x) {
        # the esf for this grouping
        esf <- meteESF(spp = x$species_code, 
                       abund = x$abundance, 
                       power = x$metabolic_rate)
        
        # sad and ipd
        s <- sad(esf)
        p <- ipd(esf)
        
        data.frame(
            # sample info
            site = x$site[1],
            site_name = x$site_name[1],
            tree = x$tree[1],
            flow_age_my = x$flow_age_my[1],
            
            # z2 values
            z2_sad = logLikZ(s)$z,
            z2_ipd = logLikZ(p)$z,
            
            # prop non-native for abundance and power
            prop_n_non_nat = sum(
                x$abundance[!(x$origin %in% c("endemic", "indigneous"))]
            ) / 
                sum(x$abundance),
            prop_p_non_nat = sum(
                x$metabolic_rate[!(x$origin %in% c("endemic", "indigneous"))]
            ) / 
                sum(x$metabolic_rate),
            
            # differences in extreme values
            n1_diff = mean(meteDist2Rank(s) == 1) - mean(s$data == 1),
            nmax_diff = sum(meteDist2Rank(s)[1:3]) - sum(s$data[1:3]),
            p1_diff = mean(meteDist2Rank(p) <= 2) - mean(p$data <= 2),
            pmax_diff = sum(meteDist2Rank(p)[1:3]) - sum(p$data[1:3])
        )
    })

arth_mete <- do.call(rbind, arth_mete)




```

An initial investigation of patterns in the SAD reveal two clear outliers in terms of deviation from theory and proportion of non-native individuals (@fig-z2-out). 


### Investigating highly invaded, highly dominated samples

The two outlier samples both show a difference (theory - observed) in maximum abundance of over -300 (@fig-z2-out). 

```{r}
outliers <- filter(arth_mete, nmax_diff < -300)
```

```{r}
#| label: fig-z2-out
#| fig-width: 7

pivot_longer(outliers, 
             c(prop_n_non_nat, nmax_diff), 
             names_to = "metric")

select(arth_mete, site_name, prop_n_non_nat, nmax_diff, z2_sad) |> 
    # pivot so both metrics are plotted in faceted panels
    pivot_longer(prop_n_non_nat:nmax_diff, names_to = "metric") |> 
    ggplot(aes(value, z2_sad, color = site_name)) +
    geom_point() +
    
    # add arrows pointing to outliers
    geom_segment(
        data = pivot_longer(outliers, 
                            c(prop_n_non_nat, nmax_diff), 
                            names_to = "metric"), 
        aes(x = value * 0.8, y = z2_sad * 0.9, 
            xend = value * 0.97, yend = z2_sad * 0.99), 
        arrow = arrow(length = unit(1, "char")),
        inherit.aes = FALSE
    ) +
    
    # facet by metric and trick strip labels into being
    # axis labels
    facet_grid(
        cols = vars(metric), scales = "free_x", 
        labeller = as_labeller(
            c(nmax_diff = "Diff. in maximum abundance", 
              prop_n_non_nat = "Prop. non-native individuals")
        ), 
        switch = "x"
    ) +
    ylab(expression("SAD"~z^2*"-value")) +
    this_ms_look() +
    theme(strip.background = element_blank(), 
          strip.placement = "outside", 
          panel.border = element_rect(colour = "black", 
                                      fill = NA, 
                                      linewidth = 1), 
          axis.title.x = element_blank())
```

These two outlier samples come from the Laupāhoehoe and Kohala sites. To further understand these two samples and determine if they should remain in the analyses we look at their composition. @tbl-out1 shows the three most abundant species for the Laupāhoehoe outlier sample while @tbl-out2 shows the three most abundant species for the `r Kohala outlier sample.


```{r}
#| label: tbl-out1

filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(
        caption = paste(
            "Top three most abundant taxa from", 
            levels(arth$site_name)[as.numeric(outliers$site[1])], 
            "outlier sample"
        )
    )

```

```{r}
#| label: tbl-out2

filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(
        caption = paste(
            "Top three most abundant taxa from", 
            levels(arth$site_name)[as.numeric(outliers$site[2])], 
            "outlier sample"
        )
    )

```

In both outliers the most abundant species is non-native, and in the case of Laupāhoehoe, all top three most abundant species are non-native. Laupāhoehoe and Kohala share *Salina celebensis* in common which is a Collembola known to reach high densities in Hawaiʻi [@christiansen1992]. The abundances of *Salina celebensis* and *Greenidea formosana* in these samples are significant outliers compared to the maximum abundance found in any other sample as shown in @fig-crazy-abund.


```{r}
#| label: fig-crazy-abund

crazy_abund <- filter(arth, 
                      (site %in% outliers$site & tree %in% outliers$tree)) |> 
    summarize(n = n(), .by = c(site, tree, species_code)) |> 
    arrange(desc(n)) |> 
    head(3)

summarize(arth, .by = c(site, tree), 
          nmax = max(table(species_code))) |> 
    filter(nmax < 300) |> 
    ggplot(aes(nmax)) +
    geom_histogram(bins = 20) +
    geom_segment(data = crazy_abund, 
                 mapping = aes(x = n, y = rep(10, 3), 
                               xend = n, yend = rep(0, 3)), 
                 color = "red", 
                 arrow = arrow(length = unit(1, "char")),
                 inherit.aes = FALSE) +
    xlab("Maximum abundances") +
    ylab("Count") +
    this_ms_look()




```

Because these samples are so unusual we decided to drop them from all further analyses presented in the main text. However, we retain the outlier samples in the supplement to validate that our results are not biased by their removal.

```{r}
#| label: remove-outliers

# remove from mete results
arth_mete_out <- arth_mete
arth_mete <- filter(arth_mete, 
                    !((site %in% outliers$site) & (tree %in% outliers$tree)))

# remove from arth data frame
arth_out <- arth
arth <- filter(arth, 
               !((site %in% outliers$site) & (tree %in% outliers$tree)))

```




## Deviations from METE across the chronosequence

Now we analyze how deviations from METE behave across the chronosequence. The following code produces the analysis of a hump-shaped versus flat relationship between METE deviation and flow age.

```{r}
#| label: z2-chrono-lm

# site-level means and sd for meta regression
site_summ <- summarize(arth_mete, 
                       site_log_age = first(log(flow_age_my)),
                       sad_m = mean(log(z2_sad)), 
                       sad_sd = sd(log(z2_sad)),
                       ipd_m = mean(log(z2_ipd)), 
                       ipd_sd = sd(log(z2_ipd)),
                       n = n(),
                       .by = site)

# full model
sad_rma_chrono <- rma(yi = sad_m, 
                      vi = sad_sd^2 / n, 
                      mods = ~ site_log_age + I(site_log_age^2), 
                      data = site_summ, 
                      method = "FE")
# reduced model (for likelihood ratio test)
sad_rma_0 <- rma(yi = sad_m, 
                 vi = sad_sd^2 / n, 
                 mods = ~ 1, 
                 data = site_summ, 
                 method = "FE")


# full, reduced as with SAD
ipd_rma_chrono <- rma(yi = ipd_m, 
                      vi = ipd_sd^2 / n, 
                      mods = ~ site_log_age + I(site_log_age^2), 
                      data = site_summ, 
                      method = "FE")

ipd_rma_0 <- rma(yi = ipd_m, 
                 vi = ipd_sd^2 / n, 
                 mods = ~ 1, 
                 data = site_summ, 
                 method = "FE")


```

Likelihood ratio tests (LRT) can evaluate the significance of the quadratic model


```{r}
#| label: z2-chrono-lrt

# likelihood ratio test for sad across chrono
sad_lrt_chrono <- anova(sad_rma_0, sad_rma_chrono)
sad_lrt_chrono

ipd_lrt_chrono <- anova(ipd_rma_0, ipd_rma_chrono)
ipd_lrt_chrono
```

The quadratic model for SAD $z^2$-values is supported by the LRT with a $P$-value < 0.0001, but the quadratic model for IPD $z^2$-values did not have a significant $P$-value. However, the 95% CI for the quadratic term in the IPD $z^2$-values model has the majority of its range below zero, indicating the data do tend toward a quadratic shape, just with greater variance compared to the quadratic model for SAD $z^2$-values.

```{r}
ipd_rma_chrono

```


The following code produces the visualization reported in the main text of deviation from METE across flow ages.

```{r}
#| label: fig-z2-chrono
#| fig-height: 7
#| fig-cap: |
#|     Deviations from METE as measured by $z^2$ across the chronosequence. 
#|     Discussed in more detail in the main text.


# prepare data frame of predicted values for plotting
x <- seq(min(site_summ$site_log_age), 
         max(site_summ$site_log_age), 
         length.out = 50)

newx <- cbind(1, x, x^2)
sad_hat <- predict(sad_rma_chrono, newmods = newx)
ipd_hat <- predict(ipd_rma_chrono, newmods = newx)

z2_chrono_pred <- rbind(
    data.frame(flow_age_my = exp(x), 
               z2 = exp(sad_hat$pred), 
               z2_lo = exp(sad_hat$ci.lb), 
               z2_hi = exp(sad_hat$ci.ub), 
               metric = "z2_sad"), 
    data.frame(flow_age_my = exp(x), 
               z2 = exp(ipd_hat$pred), 
               z2_lo = exp(ipd_hat$ci.lb), 
               z2_hi = exp(ipd_hat$ci.ub), 
               metric = "z2_ipd")
)

# plotting
fig_z2_chrono <- pivot_longer(arth_mete, c(z2_sad, z2_ipd), 
             names_to = "metric", values_to = "z2") |> 
    ggplot(aes(flow_age_my, z2, color = site_name)) +
    geom_jitter(width = 0.1) +
    # confidence envelope 
    geom_ribbon(data = z2_chrono_pred, 
                mapping = aes(x = flow_age_my, 
                              ymin = z2_lo, 
                              ymax = z2_hi), 
                color = "gray", linewidth = 0.0001, alpha = 0.25) +
    # central tendency 
    geom_line(data = z2_chrono_pred, 
              mapping = aes(flow_age_my, z2), 
              color = "black") +
    # facet by IPD or SAD and trick strip label into axis label
    facet_grid(
        rows = vars(metric), 
        labeller = as_labeller(
            c(z2_ipd = "\"IPD\"~z^2*\"-value\"", 
              z2_sad = "\"SAD\"~z^2*\"-value\""),
            label_parsed
        ), 
        switch = "y"
    ) +
    this_ms_look(log_x = TRUE, log_y = TRUE) +
    theme(axis.title.y = element_blank(),
          strip.background = element_blank(), 
          strip.placement = "outside", 
          panel.border = element_rect(colour = "black", 
                                      fill = NA, 
                                      linewidth = 1))

fig_z2_chrono

```




## How do observed patterns deviate from METE?

@fig-sad-how shows that when METE predictions deviate from observed SADs, METE under-predicts both rarity and dominance compared to observed SADs. This means that observed SADs are more uneven than the METE prediction. 


```{r}

ggplot(arth_mete, aes(z2_sad, z2_ipd, color = site)) +
    geom_point() +
    scale_x_log10(limits = range(select(arth_mete, starts_with("z2")))) +
    scale_y_log10(limits = range(select(arth_mete, starts_with("z2")))) +
    scale_color_viridis_d(option = "magma") +
    geom_abline(slope = 1, intercept = 0)


ggplot(arth_mete, aes(z2_sad, n1_diff, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(nmax_diff, z2_sad, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(p1_diff, z2_ipd, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(pmax_diff, z2_ipd, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma") 



```

## How do METE and the real data differ

Which one over-predicts rarity?

## Comparing deviations from METE with potential explanatory variables



### Deviation vs. proportion that are invasive

```{r}
#| label: fig-dev-non


ggplot(arth_mete, aes(prop_n_non_nat, z2_sad, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma")


ggplot(arth_mete, aes(prop_p_non_nat, z2_ipd, color = site)) +
    geom_point() +
    scale_y_log10() +
    scale_color_viridis_d(option = "magma")




```



### Deviation vs. dissimilarity

```{r}
#| eval: false

z_diss <- function(x, B = 500) {
    obs_dist <- x |> vegdist() |> mean()
    
    sim <- nullmodel(x, "quasiswap_count") |>
        simulate(nsim = B)

    sim_mean_dist <- sapply(1:dim(sim)[3], function(i) {
        vegdist(sim[, , i]) |> mean()
    })


    (obs_dist - mean(sim_mean_dist)) / sd(sim_mean_dist)
    
}


diss <- split(arth, arth$site) |> 
    sapply(FUN = function(dat) {
        summarize(dat, abund = n(), .by = c(tree, species_code)) |> 
            pivot_wider(names_from = species_code, 
                        values_from = abund, 
                        values_fill = 0) |> 
            select(!tree) |> 
            z_diss() 
    })

mete_by_diss <- mutate(arth_mete, 
                       mean_diss = diss[site])



ggplot(mete_by_diss, aes(mean_diss, z2_sad, color = site)) +
    geom_jitter(width = 0.25) +
    this_ms_look(log_y = TRUE)


xy <- mete_by_diss |> 
    summarize(z2_sad_mean = mean(log(z2_sad)), 
              z2_sad_sd = sd(log(z2_sad)), 
              n = n(), 
              mean_diss = first(mean_diss), 
              .by = site)



rma(yi = z2_sad_mean, 
    vi = z2_sad_sd^2 / n, 
    mods = ~ mean_diss, 
    data = xy, 
    method = "REML")


```


