---
title: "Eco-evolutionary deviations from steady state across the Hawaiian chronosequence from a maximum entropy perspective"
format: pdf
---


## Introduction

Biodiversity patterns are 

universal patterns
non-equilibrium
eco-evo and Hawaii chrono
use METE to see how systems approach or depart from non-equilib


## Methods

## Results

- deviations for SAD peak at middle age
- deviations are from more unevenness (more singletons and more hyper abundnat)
- outlier is from too much abundance of most abundant spp
- deviations not predicted by non-natives
- deviations *are* predicted by dissimilarity
- deviations for IPD less clear
    - but generally correlated with SAD
    - bigger deviations overall

## Discussion

- dissimilarity as instability
- possible causes of instability---adaptive radiation
- presents future hypothesis to test with genetic data


## Setup

```{r}
library(meteR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vegan)

# load data 
load("../data/arth.rda")

# create column for abundance, each row is an individual, therefore abundance = 1
arth$abundance <- 1

# create column for metabolic rates, assuming 3/4 scaling 
arth$metabolic_rate <- arth$individual_biomass_g^0.75

# make sites factors with levels following chronosequence
arth$site <- factor(arth$site, levels = c("KA", "MO", "KH", "LA", "VO"))
```


## METE calculations


```{r}
#| label: mete-calc
#| cache: true

arth_mete <- split(arth, arth[, c("site", "tree")], drop = TRUE) |> 
    lapply(function(x) {
        cat("\n", "site: ", x$site[1], ", tree:", x$tree[1], "\n", sep = "")
        # the esf for this grouping
        esf <- meteESF(spp = x$species_code, 
                       abund = x$abundance, 
                       power = x$metabolic_rate)
        
        # sad and ipd
        s <- sad(esf)
        p <- ipd(esf)
        
        data.frame(
            site = x$site[1],
            tree = x$tree[1],
            
            # z2 values
            z2_sad = logLikZ(s)$z,
            z2_ipd = logLikZ(p)$z,
            
            # prop non-native for abundance and power
            prop_n_non_nat = sum(
                x$abundance[!(origin %in% c("endemic", "indigneous"))]
            ) / 
                sum(x$abundance),
            prop_p_non_nat = sum(
                x$metabolic_rate[!(origin %in% c("endemic", "indigneous"))]
            ) / 
                sum(x$metabolic_rate),
            
            # differences in extreme values
            n1_ratio = mean(meteDist2Rank(s) == 1) - mean(s$data == 1),
            nmax_ratio = sum(meteDist2Rank(s)[1:3]) - sum(s$data[1:3]),
            p1_ratio = mean(meteDist2Rank(p) <= 2) - mean(p$data <= 2),
            pmax_ratio = sum(meteDist2Rank(p)[1:3]) - sum(p$data[1:3])
        )
    })

arth_mete <- do.call(rbind, arth_mete)




```


### What are the highly invaded, highly dominated samples?

```{r}
outliers <- filter(arth_mete, nmax_ratio < -300)

```

Two samples stand out as clear outliers in terms of deviation from SAD predictions as well as being the two most highly invaded samples. The two samples come from the Laupāhoehoe and Kohala sites. To further understand these two samples and determine if they should remain in the analyses we look at their composition. 


```{r}
#| label: tbl-out1


filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3) |> 
    knitr::kable(caption = paste("Top three most abundant taxa from", 
                                 outliers$tree, 
                                 "outlier sample"))

```


```{r}
outliers <- filter(arth_mete, nmax_ratio < -300)

arth_mete_out <- arth_mete
arth_mete <- filter(arth_mete, 
                    !((site %in% outliers$site) & (tree %in% outliers$tree)))

filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head(3)


filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(tray, genus, species, origin)) |> 
    arrange(tray) |> 
    tidyr::pivot_wider(names_from = tray, values_from = n) |> 
    View()


filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(tray, genus, species, origin)) |> 
    arrange(tray) |> 
    tidyr::pivot_wider(names_from = tray, values_from = n) |> 
    View()

filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head()

filter(arth, genus == "Salina" & species == "celebensis") |> 
    summarize(n = n(), .by = c(site, tree, tray)) |> 
    arrange(desc(n)) |> 
    head()


```


## Visualizing deviations from METE across chronosequence

```{r}
ggplot(arth_mete, aes(site, z2_sad, color = site)) +
    geom_jitter(width = 0.1) +
    scale_y_log10() + 
    scale_color_viridis_d(option = "magma") +
    geom_hline(yintercept = qchisq(0.95, 1))


ggplot(arth_mete, aes(site, z2_ipd, color = site)) +
    geom_jitter(width = 0.1) +
    scale_y_log10() + 
    scale_color_viridis_d(option = "magma") +
    geom_hline(yintercept = qchisq(0.95, 1))


ggplot(arth_mete, aes(site, n1_ratio, color = site)) +
    geom_jitter(width = 0.1) +
    scale_color_viridis_d(option = "magma")


ggplot(arth_mete, aes(site, nmax_ratio, color = site)) +
    geom_jitter(width = 0.1) +
    scale_color_viridis_d(option = "magma") +
    ylim(c(-150, 10))


ggplot(arth_mete, aes(z2_sad, z2_ipd, color = site)) +
    geom_point() +
    scale_x_log10(limits = range(select(arth_mete, starts_with("z2")))) +
    scale_y_log10(limits = range(select(arth_mete, starts_with("z2")))) +
    scale_color_viridis_d(option = "magma") +
    geom_abline(slope = 1, intercept = 0)


ggplot(arth_mete, aes(z2_sad, n1_ratio, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(z2_sad, nmax_ratio, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(z2_ipd, p1_ratio, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma") 

ggplot(arth_mete, aes(z2_ipd, pmax_ratio, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma") 



```

## How do METE and the real data differ

Which one over-predicts rarity?

## Comparing deviations from METE with potential explanatory variables



### Deviation vs. proportion that are invasive

```{r}
#| label: dev-inv

arth_inv <- group_by(arth, site, tree) |> 
    summarize(
        spp_prop_inv = n_distinct(
            species_code[origin %in% c("adventive", "intentional introduction")]
        ) / 
            n_distinct(species_code),
        ind_prop_inv = mean(origin %in% c("adventive", "intentional introduction")), 
        met_prop_inv = sum(
            metabolic_rate[origin %in% c("adventive", "intentional introduction")]
        ) / 
            sum(metabolic_rate)
    ) |> 
    ungroup()

mete_by_inv <- left_join(arth_mete, arth_inv, by = c("site", "tree"))


ggplot(mete_by_inv, aes(ind_prop_inv, z2_sad, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma")


ggplot(mete_by_inv, aes(ind_prop_inv, nmax_ratio, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma")



ggplot(mete_by_inv, aes(met_prop_inv, z2_ipd, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma")




```

### What are the highly invaded, highly dominated samples?

```{r}
outliers <- filter(arth_mete, nmax_ratio < -300)

filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head()


filter(arth, tree == outliers$tree[1] & site == outliers$site[1]) |> 
    summarize(n = n(), .by = c(tray, genus, species, origin)) |> 
    arrange(tray) |> 
    tidyr::pivot_wider(names_from = tray, values_from = n) |> 
    View()


filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(tray, genus, species, origin)) |> 
    arrange(tray) |> 
    tidyr::pivot_wider(names_from = tray, values_from = n) |> 
    View()

filter(arth, tree == outliers$tree[2] & site == outliers$site[2]) |> 
    summarize(n = n(), .by = c(genus, species, origin)) |> 
    arrange(desc(n)) |> 
    head()

filter(arth, genus == "Salina" & species == "celebensis") |> 
    summarize(n = n(), .by = c(site, tree, tray)) |> 
    arrange(desc(n)) |> 
    head()

filter(arth, site == "LA", tree == 2, tray == 1) |> 
    summarize(n = n(), .by = c(genus, species, origin))

filter(arth, site == "KH", tree == 6, tray == 3) |> 
    summarize(n = n(), .by = c(genus, species, origin))

```

### Deviation vs. dissimilarity

```{r}
#| eval: false

z_diss <- function(x, B = 500) {
    obs_dist <- x |> vegdist() |> mean()
    
    sim <- nullmodel(x, "quasiswap_count") |>
        simulate(nsim = B)

    sim_mean_dist <- sapply(1:dim(sim)[3], function(i) {
        vegdist(sim[, , i]) |> mean()
    })


    (obs_dist - mean(sim_mean_dist)) / sd(sim_mean_dist)
    
}


diss <- split(arth, arth$site) |> 
    sapply(FUN = function(dat) {
        summarize(dat, abund = n(), .by = c(tree, species_code)) |> 
            pivot_wider(names_from = species_code, 
                        values_from = abund, 
                        values_fill = 0) |> 
            select(!tree) |> 
            z_diss() 
    })

mete_by_diss <- mutate(arth_mete, 
                       mean_diss = diss[site])

summarize(mete_by_diss, .by = site, 
          mean_diss = first(mean_diss)) |> 
ggplot(aes(site, mean_diss, color = site)) +
    geom_point() +
    scale_color_viridis_d(option = "magma")


ggplot(mete_by_diss, aes(mean_diss, z2_sad, color = site)) +
    geom_jitter(width = 0.25) +
    scale_color_viridis_d(option = "magma")

```


